---
title: "DATS 6101 Spring 2020 Olympic Data EDA"
author: "Izzy Illari"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: readable
    toc: true
    toc_float: yes
    toc_depth: '2'
    number_sections: false
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = T, message = F)
options(scientific=T, digits = 3) 
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
# unload/detact package when done using it
```

```{r data, echo=F}
#Data taken from https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results
athlete_events <- data.frame(read.csv("/Users/iillari/Documents/gwu/spring2020/intro_dats/eda_project/120-years-of-olympic-history-athletes-and-results/athlete_events.csv"))
noc_regions <- data.frame(read.csv("/Users/iillari/Documents/gwu/spring2020/intro_dats/eda_project/120-years-of-olympic-history-athletes-and-results/noc_regions.csv"))
```

# Summary of data

Data from athlete_events.csv: 

```{r sum_athlete, echo=F}
summary(athlete_events)
```

Data from noc_regions.csv: 

```{r sum_noc, echo=F}
summary(noc_regions)
```

# Year data

Here I have manually defined a few years of data.

```{r man_year, echo=F}
#wrangle data
athlete_events_noNA <- na.omit(athlete_events)
year_data <- na.omit(athlete_events$Year)
#manually get data by year
data_2016 <- athlete_events_noNA[ athlete_events_noNA$Year == 2016, ]
data_2006 <- athlete_events_noNA[ athlete_events_noNA$Year == 2006, ]
data_1996 <- athlete_events_noNA[ athlete_events_noNA$Year == 1996, ]
#get rgb color data using console
#col2rgb(c("lightblue", "lightgreen", "pink"))
#transparent colors
light_blue <- rgb(173,216,230,max = 255, alpha = 80, names = "lt.blue")
pink <- rgb(255,192,203, max = 255, alpha = 80, names = "lt.pink")
light_green <- rgb(144,238,144, max = 255, alpha = 80, names = "lt.green")
#overlapping histograms
hist(data_2016$Weight, main= "Histogram of Olympic athlete's weights [kg] over time", xlab="weight [kg]", col = pink)
hist(data_2006$Weight, col = light_green, add=T)
hist(data_1996$Weight, col = light_blue, add=T)
legend("topright", c("2016", "2006", "1996"), fill=c(pink, light_green, light_blue))
box()
```

Here I have split up the original data by putting the data by year into a list.

```{r list_year, echo=F}
#create a list by years
data_year <- split(athlete_events_noNA, athlete_events_noNA$Year)
#summary(data_year)
hist(data_year[[35]]$Weight, main= "Histogram of Olympic athlete's weights [kg] over time", xlab="weight [kg]", col = pink)
hist(data_year[[1]]$Weight, col = light_green, add=T)
hist(data_year[[15]]$Weight, col = light_blue, add=T)
legend("topright", c("2016", "1896", "1960"), fill=c(pink, light_green, light_blue))
box()
```

There are `r length(data_year)` years. Using my data now split by years, I was to calculate the mean and variances and compare trends over time.

```{r over_time, echo=F}
years <- names(data_year)
mean_years_weight <- list()
median_years_weight <- list()
var_years_weight <- list()
sd_years_weight <- list()
for (i in 1:length(data_year)) {
  #cat("In year", i, " the mean is:", mean(data_year[[i]]$Weight), " and the std. dev. is:", sd(data_year[[i]]$Weight), "\n")
  mean_years_weight[[i]] <- mean(data_year[[i]]$Weight)
  median_years_weight[[i]] <- median(data_year[[i]]$Weight)
  var_years_weight[[i]] <- var(data_year[[i]]$Weight)
  sd_years_weight[[i]] <- sd(data_year[[i]]$Weight)
}  
list_year <- list(1896, 1900, 1904, 1906, 1908, 1912, 1920, 1924, 1928, 1932, 1936, 1948, 1952, 1956, 1960, 1964, 1968, 1972, 1976, 1980, 1984, 1988, 1992, 1994, 1996, 1998, 2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016)
```

# Scatter Plot: Average Weight Over Time

```{r make_dataframe, echo=F}
year_weight <- do.call(rbind, Map(data.frame, Year = years, Mean_Weight = mean_years_weight, Median_Weight = median_years_weight, Var_Weight = var_years_weight, StdDev_Weight = sd_years_weight))
```

Here is a scatter plot made using the mean and median weights of all Olympic athletes over time.

```{r scatter_plot_years_weight, echo=F}
plot(year_weight$Year, year_weight$Mean_Weight, xlab = "Year", ylab = "Weight [kg]", main = "Scatterplot of avg. Olympic weights [kg] over time", ylim = range(c(year_weight$Mean_Weight-year_weight$StdDev_Weight, year_weight$Mean_Weight+year_weight$StdDev_Weight)), pch=20)
#points(year_weight$Year, year_weight$Mean_Weight, pch = 20, col = "blue")
points(year_weight$Year, year_weight$Median_Weight, pch = 20, col = "red")
legend("topright", c("mean", "median"), fill=c("black", "red"))
box()
```

Here is a scatter plot made using ggplot2 to include $\pm 1 \sigma$ for the error bars.

```{r tilt_tick_marks, echo=F, results="hide"}
library(ggplot2)
library(dplyr)

theme(
  # Change axis lines
  axis.line = element_line(),
  
  # Change axis ticks text labels: font color, size and face
  axis.text = element_text(),       # Change tick labels for all axes
  axis.text.x = element_text(),     # Change x axis tick labels only
  axis.text.x.top = element_text(), # x axis tick labels on top axis
  axis.text.y = element_text(),     # Change y axis tick labels only
  axis.text.y.right = element_text(),# y  axis tick labels on top axis
  
  # Change axis ticks line: font color, size, linetype and length
  axis.ticks = element_line(),      # Change ticks line fo all axes
  axis.ticks.x = element_line(),    # Change x axis ticks only
  axis.ticks.y = element_line(),    # Change y axis ticks only
  axis.ticks.length = unit(3, "pt") # Change the length of tick marks
)
```

```{r try_ggplot, echo=F}
p1 <- ggplot(year_weight, aes(x= Year, y=Mean_Weight)) + geom_point(size=2) + geom_errorbar(aes(ymin = Mean_Weight-StdDev_Weight, ymax = Mean_Weight+StdDev_Weight))
p1 + theme(axis.text.x = element_text(angle = -90))
```

# Separate weight by other categories (sex, event, etc...)

```{r sep_by_gender, echo=F}
for (i in 1:length(data_year)) {
  women_data <- data_year[[i]][ which(data_year[[i]]$Sex == 'F'), ]
  men_data <- data_year[[i]][ which(data_year[[i]]$Sex == 'M'), ]
} 
head(women_data)
```



```{r, echo=F}
#have a name change by increments
d <- 5
for(i in 1:10) { 
 nam <- paste("A", i, sep = "")
 assign(nam, rnorm(3)+d)
}
```




