---
title: "R - In class assignment Wk #2"
author: "Izzy Illari"
date: "Jan 22, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo = TRUE, autodep = TRUE)
#dep_auto()
```

```{r basicfcn, include=F}
# Knit does not work with install.packages functions. Any install.packages statements in the RMD file needs to be commented out before knitting. 
# And if the RMD is sent to another person/computer that requires installing the package, the user can remove the commenting and install the package.
# Here, we use another option by introducting this convenient function, we named it loadPkg(), which uses install.packages() and require(). 
# The function require() is similar to library(), except it returns a TRUE/FALSE value.
# You can add quietly=T option to the require() function
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
```

#### Q1. Read in the file with all the baseball players height, weight and age info (BaseballHeightWeight.csv) as a dataframe and save it as baseball. Then display the structure of the dataframe.

```{r Q1, echo=F}
baseball <- data.frame(read.csv("BaseballHeightWeight.csv"))
str(baseball) #display structure of csv file (dataframe)
```

#### Q2. print out the first 5 and the last 3 rows of the dataframe

```{r Q2, echo=F}
head(baseball, 5) #show first 5 rows
tail(baseball, 3) #show last 3 rows
```


#### Q3. Change the headers/column names from Weight.pounds. to weight, from Height.inches. to height, and Age to age.

```{r Q3, echo=F}
invisible(library(tidyverse)) #need tidyverse package to use rename
invisible(
    baseball <- baseball %>%  rename(
        weight = Weight.pounds.,
        height = Height.inches.,
        age = Age
        )
    )
```

######## I have changed the last 3 column names to height, weight, and age.  

#### Q4. Print out the first two lines again to see the new column names.

```{r Q4, echo=F}
head(baseball, 2) #show first 2 lines
```


#### Q5. Find the statistics (summary) of the heights, weights, and ages of the players. Using complete sentences, give the reader these summary.

######## The 5-number summary of height is: `r summary(baseball$height)`; or
```{r Q5a, echo=F}
summary(baseball$height)
```

######## The 5-number summary of weight is: `r summary(baseball$weight)`; or
```{r Q5b, echo=F}
#summary(baseball)
summary(baseball$weight)
```

######## The 5-number summary of age is: `r summary(baseball$age)`; or
```{r Q5c, echo=F}
summary(baseball$age)
```

#### Q6. You should find an issue with the weight data with an NA value. Use the subset( ) function, select that data point (save as shyboy) and find out who the shy boy was.

```{r Q6, echo=F}
shyboy <- subset(baseball, is.na(weight)) #find NA value in weight
shyboy
#shyboy$Name
```

#### Q7. Remove shyboy from baseball, and then show the number of observations of the cleaned up dataframe baseball.

######## Before we remove shyboy, we have: `r length(baseball$weight)`
######## Removing shyboy...
```{r Q7, include=F}
baseball <- na.omit(baseball) #func na.omit() returns object w/ listwise deletion of NA vals
#baseball <- subset(baseball, weight>0)
```
######## shyboy removed... 
######## After removing shyboy we have: `r length(baseball$weight)` 

#### Q8. Make a quick boxplot of the weight of the players. Use the title( ) function immediately to add a title to the chart. (Try color #0000FF for blue.)

```{r Q8, echo=F}
boxplot(baseball$weight, main="boxplot of baseball players' weights", ylab="weight [lbs]", col='#0000FF')
```

#### Q9. Make a histogram of the weight of the players. Try color 'green' this time. The function understands the word 'green' just fine. Give it a better title than the default. Does the distribution looks normal?

```{r Q9, echo=F}
h <- hist(baseball$weight, main="Histogram of baseball players' weight", col = 'green', xlab="weight [lbs]") 
xfit <- seq(min(baseball$weight), max(baseball$weight), length = length(baseball$weight)) 
yfit <- dnorm(xfit, mean = mean(baseball$weight), sd = sd(baseball$weight)) 
yfit <- yfit * diff(h$mids[1:2]) * length(baseball$weight) 
lines(xfit, yfit, col = "black", lwd = 2)
```

######## The histogram shows a distribution that looks like a normal dist with mean `r mean(baseball$weight)`, median `r median(baseball$weight)`, and standard deviation `r sd(baseball$weight)`. When the mean is greater than the median, we know that a distribution is skewed to the right---as is the case here.  

#### Q10. Use ggplot, make a plot with weights vs heights of the players, color by Teams. Give it an appropriate title.

```{r Q10, echo=F}
loadPkg("ggplot2")
ggplot(data = baseball) + geom_point(mapping = aes(x = height, y = weight, color = Team)) + ggtitle("Scatter plot of baseball players' weigth [lbs] vs height [in]")
```

#### Q11. Make a plot with weights vs heights of the players, color by age

```{r Q11, echo=F}
ggplot(data = baseball) + geom_point(mapping = aes(x = height, y = weight, color = age)) + ggtitle("Scatter plot of baseball players' weigth [lbs] vs height [in]")
```

#### Q12. Obtain a subset of the data with only Team Washington. Using complete sentences, give the number of players in Team Washington.

```{r Q12, echo=F}
team_washington <- subset(baseball, Team == 'WAS')
```

######## The number of players on Team Washington is: `r length(team_washington$weight)`.

#### Q13. Obtain another subset with Team Washington and only players older than 28. How many are there in this veteran group?

```{r Q13, echo=F}
veterans <- subset(team_washington, age > 28)
```

######## The number of players on Team Washington that are older than 28 is: `r length(veterans$weight)`.

