---
title: "HW 3 EDA DATS 6101 Spring 2020"
author: "Izzy Illari"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    theme: readable
    toc: true
    toc_float: yes
    toc_depth: '2'
    number_sections: false
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = T, message = F)
options(scientific=T, digits = 3) 
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
# unload/detact package when done using it
```

# Question 1 
#### Import the dataset into R.

```{r Q1, echo=F}
pizza <- data.frame(read.csv("Pizza.csv"))
```

Pizza.csv has been imported. 

# Question 2
#### How many datapoints are there? (You can use the length( ) function.) Can you write down the answer using inline R code instead of using the code-block output? (Use include=False for the code-block if that is not the default setting for your RMD knitr.)

```{r Q2, echo=F}
num_rows <- NROW(pizza)
num_cols <- NCOL(pizza)
num_data <- num_rows*num_cols
```

The number of rows is `r num_rows` and the number of columns is `r num_cols`, meaning that there are `r num_data` data points. 

# Question 3
#### Look at the summary statistics of the dataset (try the summary( ) function).

```{r Q3, echo=F}
summary(pizza)
```

# Question 4
#### For sodium and calorie, make histograms, boxplots, and QQ-plots. Make sure all plots have appropriate titles, x- and y- labels, units on the axes if applicable. It is also much nicer to add some color to your charts instead of plain old black and white. For chart titles, if no appropriate title you can think of, just use y vs x. Don’t get mixed up (somehow I find up to half of the presentations have the wrong ones). It is NEVER x vs y. Always y vs x.

```{r Q4, echo=F}
par(mfrow=c(2,3)) #2 rows 3 columns
#sodium
boxplot(pizza$sodium, main="Boxplot of \namount of sodium \nper 100g in sample", ylab="sodium per 100g", col = '#0000FF')
hist(pizza$sodium, main="Histogram of \namount of sodium \nper 100g in sample", xlab="sodium per 100g", col = '#0000FF', density = 10) 
qqnorm(pizza$sodium, main="Q-Q plot of \namount of sodium \nper 100g in sample", col = '#0000FF') 
qqline(pizza$sodium, col = '#0000FF')
#calorie
boxplot(pizza$cal, main="Boxplot of \namount of calories \nper 100g in sample", ylab="calories per 100g", col = 'green')
hist(pizza$cal, main="Histogram of \namount of calories \nper 100g in sample", xlab="calories per 100g", col = 'green', density = 10) 
qqnorm(pizza$cal, main="Q-Q plot of \namount of calories \nper 100g in sample", col = 'green') 
qqline(pizza$cal, col = 'green')
```

# Question 5
#### Run the Shapiro-Wilk test on these two variables. What are the p-values of the test on these two variables? (No need to draw any conclusion at this point. We’ll learn these soon.)

```{r Q5, echo=F}
shapiro.test(pizza$sodium)
shapiro.test(pizza$cal)
```
The $p$-values from the Shapiro-Wilk test on the sodium and calorie data is the following:
$$p_{\text{sodium}} = `r shapiro.test(pizza$sodium)[2]`$$ 
$$p_{\text{calories}} = `r shapiro.test(pizza$cal)[2]`$$

# Question 6
#### Use the outlierKD function (or any modified version) to remove the outliers for the sodium variable, then run again to remove outliers of the calorie. Re-do the QQ-plots for these two variables. Do you see much improvements?

The code from Klodian Dhana has been modified. I have included two more input arguments into the function: one var for the specific title and one var for the color of the plot. I have also included the Shapiro-Wilk test for the $p$-value in the function, so that we may compare the Q-Q plots with and without outliers to each other.

```{r outlierKD, echo=F}
# By Klodian Dhana, https://www.r-bloggers.com/identify-describe-plot-and-remove-the-outliers-from-the-dataset/
outlierKD <- function(dt, var, title, colVal) { 
     var_name <- eval(substitute(var),eval(dt))
     na1 <- sum(is.na(var_name))
     m1 <- mean(var_name, na.rm = T)
     par(mfrow=c(1, 3), oma=c(0,0,3,0))
     outlier <- boxplot.stats(var_name)$out
     mo <- mean(outlier)
     var_name <- ifelse(var_name %in% outlier, NA, var_name)
     boxplot(var_name, main=paste("Boxplot without outliers", title), col = colVal, ylab = title)
     hist(var_name, main=paste("Histogram without outliers", title), xlab=title, col= colVal, density = 10)
     qqnorm(var_name, main=paste("Q-Q plot without outliers", title), col = colVal) 
     qqline(var_name, col = colVal)
     dt[as.character(substitute(var))] <- invisible(var_name)
     assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
     pval <- shapiro.test(var_name)[2]
     test <- pval > 0.05
     cat("p-value from Shapiro test is", toString(pval), "\n")
     cat("It is", test, "that the p-value > 0.05")
}
```

```{r Q6, echo=F}
outlierKD(pizza, sodium, "\nsodium per 100g sample", "#0000FF")
outlierKD(pizza, cal, "\ncalories per 100g sample", "green")
```

For both Q-Q plots the $p$-value is NOT greater than 0.05, implying that the distribution of the data are (perhaps) significantly different from normal distribution. However, the Shapiro test can sometimes be too sensitive. From a visual analysis it appears that the Q-Q plot for calories per 100g sample without outliers appears to follow the normal distribution. 


