---
title: "Olympic Data"
author: "team 010100"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:  
    rmdformats::readthedown:
      toc_float: true
      number_sections: true
      code_folding: hide
      keep_tex: yes
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75", scipen = 999, digits = 3, big.mark=",", warn = -1)
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r basicfunct, include=FALSE}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
```

# Introduction

With the 2020 Olympics beginning this July in Tokyo we felt that a relevant discussion to have would be `What makes an Olympian`? What can we say about Olympians? Have there been any general trends amongst Olympians? What does the Olympic population look like? These questions are all suited to EDA, and with these questions in mind we went to see if we could find data on Olympians that would be readily available for us to analyze. 

We were able to find a dataset called `120 years of Olympic history: athletes and results` on Kaggle over here: [https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results](https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results). This historical dataset includes all Olympic Games from Athens 1896 to Rio 2016, which was scraped from [https://www.sports-reference.com/](https://www.sports-reference.com/). This data was compiled by a group of Olympic historians and statisticians. All of these individuals are members of the International Society of Olympic Historians (ISOH) and have been working on this project since the late 1990s.

The report is organized as follows:

1. Summary of Dataset
2. others


# Summary of Dataset

The data looks like the following:

```{r input_data}
athlete_events <- data.frame(read.csv("athlete_events.csv"))
noc_regions <- data.frame(read.csv("noc_regions.csv"))
summary(athlete_events)
```

The athlete events data has `r ncol(athlete_events)` columns and `r nrow(athlete_events)` rows, for `r ncol(athlete_events)*nrow(athlete_events)` entries. In `athelete_events` aach row corresponds to an individual athlete competing in an individual Olympic event. The variables are the following:

1. ID: Unique number for each athlete
2. Name: Athlete's name
3. Sex: M or F
4. Age: Integer
5. Height: centimeters
6. Weight: kilograms
7. Team: Team name
8. NOC: National Olympic Committee 3-letter code
9. Games: Year and season
10. Year: Integer
11. Season: Summer or Winter
12. City: Host city
13. Sport
14. Event
15. Medal: Gold, Silver, Bronze, or NA

To prepare our data for EDA we dropped the Olympic event: Art Sculpting. NAs were also removed. 

```{r remove_NAs}
athlete_events_noNA <- na.omit(athlete_events)
data_sport <- split(athlete_events_noNA, athlete_events_noNA$Sport)
```


# Descriptive Stats

# Graphical Representation of Data

## Geographical Data

## Name Word Cloud

## Over Time

We wanted to see the changes in Olympians for the top 10 events (in terms of how many Olympians were recorded for the sport) over the years. Below are the top 10 events:  

```{r top_ten, echo=F}
list_sports <- as.data.frame(table(athlete_events_noNA$Sport))[1]
sport_counts <- as.data.frame(table(athlete_events_noNA$Sport))[2]
count_sports_data <- do.call(rbind, Map(data.frame, "sport.names" = list_sports, "sport.counts" = sport_counts))
sort_count_sports_data <- count_sports_data[order(-count_sports_data$sport.counts),]
sort_count_sports_data[2:11,]
top10 <- sort_count_sports_data[2:11,][1]
```

In list form:

1. Swimming
2. Rowing
3. Ice Hockey
4. Hockey
5. Gymnastics
6. Fencing
7. Football
8. Canoeing
9. Basketball
10. Wrestling

```{r over_time, echo=F}
# Swimming
data_Swimming_gender <- split(data_sport$Swimming, data_sport$Swimming$Sex)
data_Swimming_F_year <- split(data_Swimming_gender$F, data_Swimming_gender$F$Year)
data_Swimming_M_year <- split(data_Swimming_gender$"M", data_Swimming_gender$"M"$Year)
years_F_Swimming <- names(data_Swimming_F_year)
years_M_Swimming <- names(data_Swimming_M_year)
mean_years_weight_F_Swimming <- list()
sd_years_weight_F_Swimming <- list()
mean_years_height_F_Swimming <- list()
sd_years_height_F_Swimming <- list()
mean_years_weight_M_Swimming <- list()
sd_years_weight_M_Swimming <- list()
mean_years_height_M_Swimming <- list()
sd_years_height_M_Swimming <- list()
# women data
for (i in 1:length(data_Swimming_F_year)) {
	mean_years_weight_F_Swimming[[i]] <- mean(data_Swimming_F_year[[i]]$Weight)
	sd_years_weight_F_Swimming[[i]] <- sd(data_Swimming_F_year[[i]]$Weight)
	mean_years_height_F_Swimming[[i]] <- mean(data_Swimming_F_year[[i]]$Height)
	sd_years_height_F_Swimming[[i]] <- sd(data_Swimming_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Swimming_M_year)) {
	mean_years_weight_M_Swimming[[i]] <- mean(data_Swimming_M_year[[i]]$Weight)
	sd_years_weight_M_Swimming[[i]] <- sd(data_Swimming_M_year[[i]]$Weight)
	mean_years_height_M_Swimming[[i]] <- mean(data_Swimming_M_year[[i]]$Height)
	sd_years_height_M_Swimming[[i]] <- sd(data_Swimming_M_year[[i]]$Height)
}
# make data frames
Swimming_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Swimming, Mean_Weight = mean_years_weight_F_Swimming, StdDev_Weight = sd_years_weight_F_Swimming, Mean_Height = mean_years_height_F_Swimming, StdDev_Height = sd_years_height_F_Swimming))
Swimming_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Swimming, Mean_Weight = mean_years_weight_F_Swimming, StdDev_Weight = sd_years_weight_F_Swimming, Mean_Height = mean_years_height_F_Swimming, StdDev_Height = sd_years_height_F_Swimming))
# Rowing
data_Rowing_gender <- split(data_sport$Rowing, data_sport$Rowing$Sex)
data_Rowing_F_year <- split(data_Rowing_gender$F, data_Rowing_gender$F$Year)
data_Rowing_M_year <- split(data_Rowing_gender$"M", data_Rowing_gender$"M"$Year)
years_F_Rowing <- names(data_Rowing_F_year)
years_M_Rowing <- names(data_Rowing_M_year)
mean_years_weight_F_Rowing <- list()
sd_years_weight_F_Rowing <- list()
mean_years_height_F_Rowing <- list()
sd_years_height_F_Rowing <- list()
mean_years_weight_M_Rowing <- list()
sd_years_weight_M_Rowing <- list()
mean_years_height_M_Rowing <- list()
sd_years_height_M_Rowing <- list()
# women data
for (i in 1:length(data_Rowing_F_year)) {
	mean_years_weight_F_Rowing[[i]] <- mean(data_Rowing_F_year[[i]]$Weight)
	sd_years_weight_F_Rowing[[i]] <- sd(data_Rowing_F_year[[i]]$Weight)
	mean_years_height_F_Rowing[[i]] <- mean(data_Rowing_F_year[[i]]$Height)
	sd_years_height_F_Rowing[[i]] <- sd(data_Rowing_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Rowing_M_year)) {
	mean_years_weight_M_Rowing[[i]] <- mean(data_Rowing_M_year[[i]]$Weight)
	sd_years_weight_M_Rowing[[i]] <- sd(data_Rowing_M_year[[i]]$Weight)
	mean_years_height_M_Rowing[[i]] <- mean(data_Rowing_M_year[[i]]$Height)
	sd_years_height_M_Rowing[[i]] <- sd(data_Rowing_M_year[[i]]$Height)
}
# make data frames
Rowing_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Rowing, Mean_Weight = mean_years_weight_F_Rowing, StdDev_Weight = sd_years_weight_F_Rowing, Mean_Height = mean_years_height_F_Rowing, StdDev_Height = sd_years_height_F_Rowing))
Rowing_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Rowing, Mean_Weight = mean_years_weight_F_Rowing, StdDev_Weight = sd_years_weight_F_Rowing, Mean_Height = mean_years_height_F_Rowing, StdDev_Height = sd_years_height_F_Rowing))
# Ice.Hockey
data_Ice.Hockey_gender <- split(data_sport$"Ice Hockey", data_sport$"Ice Hockey"$Sex)
data_Ice.Hockey_F_year <- split(data_Ice.Hockey_gender$F, data_Ice.Hockey_gender$F$Year)
data_Ice.Hockey_M_year <- split(data_Ice.Hockey_gender$"M", data_Ice.Hockey_gender$"M"$Year)
years_F_Ice.Hockey <- names(data_Ice.Hockey_F_year)
years_M_Ice.Hockey <- names(data_Ice.Hockey_M_year)
mean_years_weight_F_Ice.Hockey <- list()
sd_years_weight_F_Ice.Hockey <- list()
mean_years_height_F_Ice.Hockey <- list()
sd_years_height_F_Ice.Hockey <- list()
mean_years_weight_M_Ice.Hockey <- list()
sd_years_weight_M_Ice.Hockey <- list()
mean_years_height_M_Ice.Hockey <- list()
sd_years_height_M_Ice.Hockey <- list()
# women data
for (i in 1:length(data_Ice.Hockey_F_year)) {
	mean_years_weight_F_Ice.Hockey[[i]] <- mean(data_Ice.Hockey_F_year[[i]]$Weight)
	sd_years_weight_F_Ice.Hockey[[i]] <- sd(data_Ice.Hockey_F_year[[i]]$Weight)
	mean_years_height_F_Ice.Hockey[[i]] <- mean(data_Ice.Hockey_F_year[[i]]$Height)
	sd_years_height_F_Ice.Hockey[[i]] <- sd(data_Ice.Hockey_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Ice.Hockey_M_year)) {
	mean_years_weight_M_Ice.Hockey[[i]] <- mean(data_Ice.Hockey_M_year[[i]]$Weight)
	sd_years_weight_M_Ice.Hockey[[i]] <- sd(data_Ice.Hockey_M_year[[i]]$Weight)
	mean_years_height_M_Ice.Hockey[[i]] <- mean(data_Ice.Hockey_M_year[[i]]$Height)
	sd_years_height_M_Ice.Hockey[[i]] <- sd(data_Ice.Hockey_M_year[[i]]$Height)
}
# make data frames
Ice.Hockey_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Ice.Hockey, Mean_Weight = mean_years_weight_F_Ice.Hockey, StdDev_Weight = sd_years_weight_F_Ice.Hockey, Mean_Height = mean_years_height_F_Ice.Hockey, StdDev_Height = sd_years_height_F_Ice.Hockey))
Ice.Hockey_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Ice.Hockey, Mean_Weight = mean_years_weight_F_Ice.Hockey, StdDev_Weight = sd_years_weight_F_Ice.Hockey, Mean_Height = mean_years_height_F_Ice.Hockey, StdDev_Height = sd_years_height_F_Ice.Hockey))
# Hockey
data_Hockey_gender <- split(data_sport$Hockey, data_sport$Hockey$Sex)
data_Hockey_F_year <- split(data_Hockey_gender$F, data_Hockey_gender$F$Year)
data_Hockey_M_year <- split(data_Hockey_gender$"M", data_Hockey_gender$"M"$Year)
years_F_Hockey <- names(data_Hockey_F_year)
years_M_Hockey <- names(data_Hockey_M_year)
mean_years_weight_F_Hockey <- list()
sd_years_weight_F_Hockey <- list()
mean_years_height_F_Hockey <- list()
sd_years_height_F_Hockey <- list()
mean_years_weight_M_Hockey <- list()
sd_years_weight_M_Hockey <- list()
mean_years_height_M_Hockey <- list()
sd_years_height_M_Hockey <- list()
# women data
for (i in 1:length(data_Hockey_F_year)) {
	mean_years_weight_F_Hockey[[i]] <- mean(data_Hockey_F_year[[i]]$Weight)
	sd_years_weight_F_Hockey[[i]] <- sd(data_Hockey_F_year[[i]]$Weight)
	mean_years_height_F_Hockey[[i]] <- mean(data_Hockey_F_year[[i]]$Height)
	sd_years_height_F_Hockey[[i]] <- sd(data_Hockey_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Hockey_M_year)) {
	mean_years_weight_M_Hockey[[i]] <- mean(data_Hockey_M_year[[i]]$Weight)
	sd_years_weight_M_Hockey[[i]] <- sd(data_Hockey_M_year[[i]]$Weight)
	mean_years_height_M_Hockey[[i]] <- mean(data_Hockey_M_year[[i]]$Height)
	sd_years_height_M_Hockey[[i]] <- sd(data_Hockey_M_year[[i]]$Height)
}
# make data frames
Hockey_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Hockey, Mean_Weight = mean_years_weight_F_Hockey, StdDev_Weight = sd_years_weight_F_Hockey, Mean_Height = mean_years_height_F_Hockey, StdDev_Height = sd_years_height_F_Hockey))
Hockey_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Hockey, Mean_Weight = mean_years_weight_F_Hockey, StdDev_Weight = sd_years_weight_F_Hockey, Mean_Height = mean_years_height_F_Hockey, StdDev_Height = sd_years_height_F_Hockey))
# Gymnastics
data_Gymnastics_gender <- split(data_sport$Gymnastics, data_sport$Gymnastics$Sex)
data_Gymnastics_F_year <- split(data_Gymnastics_gender$F, data_Gymnastics_gender$F$Year)
data_Gymnastics_M_year <- split(data_Gymnastics_gender$"M", data_Gymnastics_gender$"M"$Year)
years_F_Gymnastics <- names(data_Gymnastics_F_year)
years_M_Gymnastics <- names(data_Gymnastics_M_year)
mean_years_weight_F_Gymnastics <- list()
sd_years_weight_F_Gymnastics <- list()
mean_years_height_F_Gymnastics <- list()
sd_years_height_F_Gymnastics <- list()
mean_years_weight_M_Gymnastics <- list()
sd_years_weight_M_Gymnastics <- list()
mean_years_height_M_Gymnastics <- list()
sd_years_height_M_Gymnastics <- list()
# women data
for (i in 1:length(data_Gymnastics_F_year)) {
	mean_years_weight_F_Gymnastics[[i]] <- mean(data_Gymnastics_F_year[[i]]$Weight)
	sd_years_weight_F_Gymnastics[[i]] <- sd(data_Gymnastics_F_year[[i]]$Weight)
	mean_years_height_F_Gymnastics[[i]] <- mean(data_Gymnastics_F_year[[i]]$Height)
	sd_years_height_F_Gymnastics[[i]] <- sd(data_Gymnastics_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Gymnastics_M_year)) {
	mean_years_weight_M_Gymnastics[[i]] <- mean(data_Gymnastics_M_year[[i]]$Weight)
	sd_years_weight_M_Gymnastics[[i]] <- sd(data_Gymnastics_M_year[[i]]$Weight)
	mean_years_height_M_Gymnastics[[i]] <- mean(data_Gymnastics_M_year[[i]]$Height)
	sd_years_height_M_Gymnastics[[i]] <- sd(data_Gymnastics_M_year[[i]]$Height)
}
# make data frames
Gymnastics_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Gymnastics, Mean_Weight = mean_years_weight_F_Gymnastics, StdDev_Weight = sd_years_weight_F_Gymnastics, Mean_Height = mean_years_height_F_Gymnastics, StdDev_Height = sd_years_height_F_Gymnastics))
Gymnastics_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Gymnastics, Mean_Weight = mean_years_weight_F_Gymnastics, StdDev_Weight = sd_years_weight_F_Gymnastics, Mean_Height = mean_years_height_F_Gymnastics, StdDev_Height = sd_years_height_F_Gymnastics))
# Fencing
data_Fencing_gender <- split(data_sport$Fencing, data_sport$Fencing$Sex)
data_Fencing_F_year <- split(data_Fencing_gender$F, data_Fencing_gender$F$Year)
data_Fencing_M_year <- split(data_Fencing_gender$"M", data_Fencing_gender$"M"$Year)
years_F_Fencing <- names(data_Fencing_F_year)
years_M_Fencing <- names(data_Fencing_M_year)
mean_years_weight_F_Fencing <- list()
sd_years_weight_F_Fencing <- list()
mean_years_height_F_Fencing <- list()
sd_years_height_F_Fencing <- list()
mean_years_weight_M_Fencing <- list()
sd_years_weight_M_Fencing <- list()
mean_years_height_M_Fencing <- list()
sd_years_height_M_Fencing <- list()
# women data
for (i in 1:length(data_Fencing_F_year)) {
	mean_years_weight_F_Fencing[[i]] <- mean(data_Fencing_F_year[[i]]$Weight)
	sd_years_weight_F_Fencing[[i]] <- sd(data_Fencing_F_year[[i]]$Weight)
	mean_years_height_F_Fencing[[i]] <- mean(data_Fencing_F_year[[i]]$Height)
	sd_years_height_F_Fencing[[i]] <- sd(data_Fencing_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Fencing_M_year)) {
	mean_years_weight_M_Fencing[[i]] <- mean(data_Fencing_M_year[[i]]$Weight)
	sd_years_weight_M_Fencing[[i]] <- sd(data_Fencing_M_year[[i]]$Weight)
	mean_years_height_M_Fencing[[i]] <- mean(data_Fencing_M_year[[i]]$Height)
	sd_years_height_M_Fencing[[i]] <- sd(data_Fencing_M_year[[i]]$Height)
}
# make data frames
Fencing_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Fencing, Mean_Weight = mean_years_weight_F_Fencing, StdDev_Weight = sd_years_weight_F_Fencing, Mean_Height = mean_years_height_F_Fencing, StdDev_Height = sd_years_height_F_Fencing))
Fencing_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Fencing, Mean_Weight = mean_years_weight_F_Fencing, StdDev_Weight = sd_years_weight_F_Fencing, Mean_Height = mean_years_height_F_Fencing, StdDev_Height = sd_years_height_F_Fencing))
# Football
data_Football_gender <- split(data_sport$Football, data_sport$Football$Sex)
data_Football_F_year <- split(data_Football_gender$F, data_Football_gender$F$Year)
data_Football_M_year <- split(data_Football_gender$"M", data_Football_gender$"M"$Year)
years_F_Football <- names(data_Football_F_year)
years_M_Football <- names(data_Football_M_year)
mean_years_weight_F_Football <- list()
sd_years_weight_F_Football <- list()
mean_years_height_F_Football <- list()
sd_years_height_F_Football <- list()
mean_years_weight_M_Football <- list()
sd_years_weight_M_Football <- list()
mean_years_height_M_Football <- list()
sd_years_height_M_Football <- list()
# women data
for (i in 1:length(data_Football_F_year)) {
	mean_years_weight_F_Football[[i]] <- mean(data_Football_F_year[[i]]$Weight)
	sd_years_weight_F_Football[[i]] <- sd(data_Football_F_year[[i]]$Weight)
	mean_years_height_F_Football[[i]] <- mean(data_Football_F_year[[i]]$Height)
	sd_years_height_F_Football[[i]] <- sd(data_Football_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Football_M_year)) {
	mean_years_weight_M_Football[[i]] <- mean(data_Football_M_year[[i]]$Weight)
	sd_years_weight_M_Football[[i]] <- sd(data_Football_M_year[[i]]$Weight)
	mean_years_height_M_Football[[i]] <- mean(data_Football_M_year[[i]]$Height)
	sd_years_height_M_Football[[i]] <- sd(data_Football_M_year[[i]]$Height)
}
# make data frames
Football_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Football, Mean_Weight = mean_years_weight_F_Football, StdDev_Weight = sd_years_weight_F_Football, Mean_Height = mean_years_height_F_Football, StdDev_Height = sd_years_height_F_Football))
Football_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Football, Mean_Weight = mean_years_weight_F_Football, StdDev_Weight = sd_years_weight_F_Football, Mean_Height = mean_years_height_F_Football, StdDev_Height = sd_years_height_F_Football))
# Canoeing
data_Canoeing_gender <- split(data_sport$Canoeing, data_sport$Canoeing$Sex)
data_Canoeing_F_year <- split(data_Canoeing_gender$F, data_Canoeing_gender$F$Year)
data_Canoeing_M_year <- split(data_Canoeing_gender$"M", data_Canoeing_gender$"M"$Year)
years_F_Canoeing <- names(data_Canoeing_F_year)
years_M_Canoeing <- names(data_Canoeing_M_year)
mean_years_weight_F_Canoeing <- list()
sd_years_weight_F_Canoeing <- list()
mean_years_height_F_Canoeing <- list()
sd_years_height_F_Canoeing <- list()
mean_years_weight_M_Canoeing <- list()
sd_years_weight_M_Canoeing <- list()
mean_years_height_M_Canoeing <- list()
sd_years_height_M_Canoeing <- list()
# women data
for (i in 1:length(data_Canoeing_F_year)) {
	mean_years_weight_F_Canoeing[[i]] <- mean(data_Canoeing_F_year[[i]]$Weight)
	sd_years_weight_F_Canoeing[[i]] <- sd(data_Canoeing_F_year[[i]]$Weight)
	mean_years_height_F_Canoeing[[i]] <- mean(data_Canoeing_F_year[[i]]$Height)
	sd_years_height_F_Canoeing[[i]] <- sd(data_Canoeing_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Canoeing_M_year)) {
	mean_years_weight_M_Canoeing[[i]] <- mean(data_Canoeing_M_year[[i]]$Weight)
	sd_years_weight_M_Canoeing[[i]] <- sd(data_Canoeing_M_year[[i]]$Weight)
	mean_years_height_M_Canoeing[[i]] <- mean(data_Canoeing_M_year[[i]]$Height)
	sd_years_height_M_Canoeing[[i]] <- sd(data_Canoeing_M_year[[i]]$Height)
}
# make data frames
Canoeing_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Canoeing, Mean_Weight = mean_years_weight_F_Canoeing, StdDev_Weight = sd_years_weight_F_Canoeing, Mean_Height = mean_years_height_F_Canoeing, StdDev_Height = sd_years_height_F_Canoeing))
Canoeing_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Canoeing, Mean_Weight = mean_years_weight_F_Canoeing, StdDev_Weight = sd_years_weight_F_Canoeing, Mean_Height = mean_years_height_F_Canoeing, StdDev_Height = sd_years_height_F_Canoeing))
# Basketball
data_Basketball_gender <- split(data_sport$Basketball, data_sport$Basketball$Sex)
data_Basketball_F_year <- split(data_Basketball_gender$F, data_Basketball_gender$F$Year)
data_Basketball_M_year <- split(data_Basketball_gender$"M", data_Basketball_gender$"M"$Year)
years_F_Basketball <- names(data_Basketball_F_year)
years_M_Basketball <- names(data_Basketball_M_year)
mean_years_weight_F_Basketball <- list()
sd_years_weight_F_Basketball <- list()
mean_years_height_F_Basketball <- list()
sd_years_height_F_Basketball <- list()
mean_years_weight_M_Basketball <- list()
sd_years_weight_M_Basketball <- list()
mean_years_height_M_Basketball <- list()
sd_years_height_M_Basketball <- list()
# women data
for (i in 1:length(data_Basketball_F_year)) {
	mean_years_weight_F_Basketball[[i]] <- mean(data_Basketball_F_year[[i]]$Weight)
	sd_years_weight_F_Basketball[[i]] <- sd(data_Basketball_F_year[[i]]$Weight)
	mean_years_height_F_Basketball[[i]] <- mean(data_Basketball_F_year[[i]]$Height)
	sd_years_height_F_Basketball[[i]] <- sd(data_Basketball_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Basketball_M_year)) {
	mean_years_weight_M_Basketball[[i]] <- mean(data_Basketball_M_year[[i]]$Weight)
	sd_years_weight_M_Basketball[[i]] <- sd(data_Basketball_M_year[[i]]$Weight)
	mean_years_height_M_Basketball[[i]] <- mean(data_Basketball_M_year[[i]]$Height)
	sd_years_height_M_Basketball[[i]] <- sd(data_Basketball_M_year[[i]]$Height)
}
# make data frames
Basketball_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Basketball, Mean_Weight = mean_years_weight_F_Basketball, StdDev_Weight = sd_years_weight_F_Basketball, Mean_Height = mean_years_height_F_Basketball, StdDev_Height = sd_years_height_F_Basketball))
Basketball_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Basketball, Mean_Weight = mean_years_weight_F_Basketball, StdDev_Weight = sd_years_weight_F_Basketball, Mean_Height = mean_years_height_F_Basketball, StdDev_Height = sd_years_height_F_Basketball))
# Wrestling
data_Wrestling_gender <- split(data_sport$Wrestling, data_sport$Wrestling$Sex)
data_Wrestling_F_year <- split(data_Wrestling_gender$F, data_Wrestling_gender$F$Year)
data_Wrestling_M_year <- split(data_Wrestling_gender$"M", data_Wrestling_gender$"M"$Year)
years_F_Wrestling <- names(data_Wrestling_F_year)
years_M_Wrestling <- names(data_Wrestling_M_year)
mean_years_weight_F_Wrestling <- list()
sd_years_weight_F_Wrestling <- list()
mean_years_height_F_Wrestling <- list()
sd_years_height_F_Wrestling <- list()
mean_years_weight_M_Wrestling <- list()
sd_years_weight_M_Wrestling <- list()
mean_years_height_M_Wrestling <- list()
sd_years_height_M_Wrestling <- list()
# women data
for (i in 1:length(data_Wrestling_F_year)) {
	mean_years_weight_F_Wrestling[[i]] <- mean(data_Wrestling_F_year[[i]]$Weight)
	sd_years_weight_F_Wrestling[[i]] <- sd(data_Wrestling_F_year[[i]]$Weight)
	mean_years_height_F_Wrestling[[i]] <- mean(data_Wrestling_F_year[[i]]$Height)
	sd_years_height_F_Wrestling[[i]] <- sd(data_Wrestling_F_year[[i]]$Height)
}
# men data
for (i in 1:length(data_Wrestling_M_year)) {
	mean_years_weight_M_Wrestling[[i]] <- mean(data_Wrestling_M_year[[i]]$Weight)
	sd_years_weight_M_Wrestling[[i]] <- sd(data_Wrestling_M_year[[i]]$Weight)
	mean_years_height_M_Wrestling[[i]] <- mean(data_Wrestling_M_year[[i]]$Height)
	sd_years_height_M_Wrestling[[i]] <- sd(data_Wrestling_M_year[[i]]$Height)
}
# make data frames
Wrestling_F_data <- do.call(rbind, Map(data.frame, Year = years_F_Wrestling, Mean_Weight = mean_years_weight_F_Wrestling, StdDev_Weight = sd_years_weight_F_Wrestling, Mean_Height = mean_years_height_F_Wrestling, StdDev_Height = sd_years_height_F_Wrestling))
Wrestling_M_data <- do.call(rbind, Map(data.frame, Year = years_M_Wrestling, Mean_Weight = mean_years_weight_F_Wrestling, StdDev_Weight = sd_years_weight_F_Wrestling, Mean_Height = mean_years_height_F_Wrestling, StdDev_Height = sd_years_height_F_Wrestling))
```

```{r testggplot, echo=F}
# women weight
plot_swim_F_w <- ggplot(Swimming_F_data, aes(x=Year, y=Mean_Weight, group=1)) + geom_line(color='pink') + geom_point(color='pink')
plot_swim_F_w
plot_row_F_w <- ggplot(Rowing_F_data, aes(x=Year, y=Mean_Weight, group=1)) + geom_point(colour='red') + geom_line(colour='red')

women_weight_plot <- plot_swim_F_w + geom_point(data = Rowing_F_data, colour='red') + geom_line() + geom_point(data = Ice.Hockey_F_data, colour='pink') + geom_point(data = Hockey_F_data, colour='coral') + geom_point(data = Gymnastics_F_data, colour='orange') + geom_point(data = Fencing_F_data, colour='gold') + geom_point(data = Football_F_data, colour='green') + geom_point(data = Canoeing_F_data, colour='lightseagreen') + geom_point(data = Basketball_F_data, colour='blue') + geom_point(data = Wrestling_F_data, colour='purple') + theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Scatterplot of Olympic Women's Average Weights [kg] Over Time", x="year", y="weight [kg]", color = "Legend") 
# women height
women_height_plot <- ggplot(Swimming_F_data, aes(x=Year, y=Mean_Height)) + geom_point() + labs(title="Scatterplot of Olympic Women's Average Heights [cm] Over Time", x="year", y="height [cm]") + geom_point(data = Rowing_F_data, colour='red') + geom_point(data = Ice.Hockey_F_data, colour='pink') + geom_point(data = Hockey_F_data, colour='coral') + geom_point(data = Gymnastics_F_data, colour='orange') + geom_point(data = Fencing_F_data, colour='gold') + geom_point(data = Football_F_data, colour='green') + geom_point(data = Canoeing_F_data, colour='lightseagreen') + geom_point(data = Basketball_F_data, colour='blue') + geom_point(data = Wrestling_F_data, colour='purple') + theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1))
# men weight
men_weight_plot <- ggplot(Swimming_M_data, aes(x=Year, y=Mean_Weight)) + geom_point() + labs(title="Scatterplot of Olympic Men's Average Weights [kg] Over Time", x="year", y="weight [kg]") + geom_point(data = Rowing_M_data, colour='red') + geom_point(data = Ice.Hockey_M_data, colour='pink') + geom_point(data = Hockey_M_data, colour='coral') + geom_point(data = Gymnastics_M_data, colour='orange') + geom_point(data = Fencing_M_data, colour='gold') + geom_point(data = Football_M_data, colour='green') + geom_point(data = Canoeing_M_data, colour='lightseagreen') + geom_point(data = Basketball_M_data, colour='blue') + geom_point(data = Wrestling_M_data, colour='purple') + theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1))
# men height
men_height_plot <- ggplot(Swimming_M_data, aes(x=Year, y=Mean_Height)) + geom_point() + labs(title="Scatterplot of Olympic Men's Average Heights [cm] Over Time", x="year", y="height [cm]") + geom_point(data = Rowing_M_data, colour='red') + geom_point(data = Ice.Hockey_M_data, colour='pink') + geom_point(data = Hockey_M_data, colour='coral') + geom_point(data = Gymnastics_M_data, colour='orange') + geom_point(data = Fencing_M_data, colour='gold') + geom_point(data = Football_M_data, colour='green') + geom_point(data = Canoeing_M_data, colour='lightseagreen') + geom_point(data = Basketball_M_data, colour='blue') + geom_point(data = Wrestling_M_data, colour='purple') + theme_minimal() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1))
# plots
women_weight_plot
women_height_plot
men_weight_plot
men_height_plot
		

# Variance/Standard Deviation/Histograms

# Normality Tests
