---
title: "Intro to DS - HW04"
author: "Izzy Illari"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    keep_tex: yes
    number_sections: no
    theme: readable
    toc: yes
    toc_depth: 2
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, results = T, message = F)
options(scientific=T, digits = 3) 
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
detach_package = function(pkg, character.only = FALSE) { if(!character.only) { pkg <- deparse(substitute(pkg)) } 
  search_item <- paste("package", pkg,sep = ":") 
  while(search_item %in% search()) { detach(search_item, unload = TRUE, character.only = TRUE) } 
}
```


# Question 1

**After importing the dataset as Adata, check the structure. Make sure admit and rank are stored as factors, instead of numeric or integers. (Hint: Use as.factor() function or factor() function).**

Importing LogReg.csv...

```{r q01a, results=F}
Adata <- data.frame(read.csv("LogReg.csv"))
is.factor(Adata$admit)
is.factor(Adata$rank)
```

First I check if admit and rank are factors by using is.factor(). It is `r is.factor(Adata$admit)` that admit is a factor. It is `r is.factor(Adata$rank)` that rank is a factor. Since neither are factors, I use factor() to coerce the argument to a factor.

```{r q01b, results=F}
Adata$admit <- factor(Adata$admit)
Adata$rank <- factor(Adata$rank)
is.factor(Adata$admit)
is.factor(Adata$rank)
```

Now I check if I have successfully coerced the arguments of admit and rank to a factor. It is `r is.factor(Adata$admit)` that admit is a factor. It is `r is.factor(Adata$rank)` that rank is a factor.  


# Question 2
**Construct t-intervals for the gre and gpa data for all applicants at 0.80 level and 0.99 level. **

First I construct the t-intervals for the gre data for all applicants at 0.80 level and 0.99 level.

```{r q2a, results=T}
ttest_gre_all_80 = t.test(x=Adata$gre, conf.level=0.80 )
ttest_gre_all_99 = t.test(x=Adata$gre, conf.level=0.99 )
#ttest_gre_all_80
```

I check the confidence intervals to see if it has been set correctly. I have the following: 

```{r q2a1, results=T}
#names(ttest_gre_all_80)
#ttest_gre_all_80$alternative
#ttest_gre_all_80$estimate
ttest_gre_all_80$conf.int
ttest_gre_all_99$conf.int
```

Then I construct the t-intervals for the gpa data for all applicants at 0.80 level and 0.99 level.

```{r q2b, results=T}
ttest_gpa_all_80 = t.test(x=Adata$gpa, conf.level=0.80 )
ttest_gpa_all_99 = t.test(x=Adata$gpa, conf.level=0.99 )
```

I check the confidence intervals to see if it has been set correctly. I have the following: 

```{r q2b1, results=T}
ttest_gpa_all_80$conf.int
ttest_gpa_all_99$conf.int
```


# Question 3
**Repeat the same calculation for gre in Question 2 but for admitted (1) and rejected (0) separately. (You can subset them first.) Between the admitted and rejected, does the two intervals overlap at 0.80 level? And at 0.99 level?**

I will subset the gre data by testing if admit is 1 or 0, and then set 1 for admitted and 0 for rejected. Then I will construct t-intervals at the 0.80 and 0.99 levels for the admitted and rejected data. 

```{r q3a}
#subset data by admit or reject
gre_admit <- subset(Adata, admit == 1, select = c("admit","gre","gpa","rank"))
gre_reject <- subset(Adata, admit == 0, select = c("admit","gre","gpa","rank"))
#t test
ttest_gre_admit_80 = t.test(x=gre_admit$gre, conf.level=0.80 )
ttest_gre_admit_99 = t.test(x=gre_admit$gre, conf.level=0.99 )
ttest_gre_reject_80 = t.test(x=gre_reject$gre, conf.level=0.80 )
ttest_gre_reject_99 = t.test(x=gre_reject$gre, conf.level=0.99 )
```

After subsetting the data we want to check if the admitted and rejected CIs overlap. In order to check if intervals overlap it means there exists some number $c$ which is in both ranges, i.e. 

$$
x_1 \leq c \leq x_2
$$

and 

$$
y_1 \leq c \leq y_2
$$
If we assume that the ranges are well-formed (such that $x_1 \leq x_2$ and $y_1 \leq y_2$) then it is sufficient to test

$$
x_1 \leq y_2 \text{ and } y_1 \leq x_2
$$

At the 0.80 level the CI for rejected is `r ttest_gre_reject_80$conf.int` and the CI for admitted is `r ttest_gre_admit_80$conf.int`. Then we check $x_{r} \leq y_{a}$, which gives us `r ttest_gre_reject_80$conf.int[1] <= ttest_gre_admit_80$conf.int[2]`, and then we check $y_{r} \leq x_{a}$, which gives us `r ttest_gre_reject_80$conf.int[2] <= ttest_gre_admit_80$conf.int[1]`. Because both tests return TRUE, this means that at the 0.80 the CIs do not overlap.   

Now we check if the admitted and rejected overlap at the 0.99 level in the same way. At the 0.99 level the CI for rejected is `r ttest_gre_reject_99$conf.int` and the CI for admitted is `r ttest_gre_admit_99$conf.int`. Then we check $x_{r} \leq y_{a}$, which gives us `r ttest_gre_reject_99$conf.int[1] <= ttest_gre_admit_99$conf.int[2]`, and then we check $y_{r} \leq x_{a}$, which gives us `r ttest_gre_reject_99$conf.int[2] <= ttest_gre_admit_99$conf.int[1]`. Because both tests return TRUE, this means that at the 0.99 the CIs do not overlap. 


# Question 4
**Make (box-) plots showing the gre distribution among applicants from different school rankings for the admitted and rejected separately. Please use ggplot for this. The x-variable should be rank, and the y-variable should be gre.**

```{r q4, results=T}
library(ggplot2)
library(gridExtra)
#admitted
p_gre_admit <- ggplot(gre_admit, aes(x=gre_admit$rank, y=gre_admit$gre)) + geom_boxplot() + stat_summary(fun.y=mean, geom="point", shape=23, size=4) + labs(title="Boxplot of gre scores \nvs school rankings \nof admitted students", x="school rank", y="gre scores")
#rejected
p_gre_reject <- ggplot(gre_reject, aes(x=gre_reject$rank, y=gre_reject$gre)) + geom_boxplot() + stat_summary(fun.y=mean, geom="point", shape=23, size=4) + labs(title="Boxplot of gre scores \nvs school rankings \nof rejected students", x="school rank", y="gre scores")
grid.arrange(p_gre_admit, p_gre_reject, ncol=2)
```


# Question 5
**Repeat Question 3 for gpa. Do the two groups have t-intervals overlap at 0.80 level? At 0.99 level?**

I have already subsetted the data by admitted or rejected in Question 3. I will construct the t-intervals for the gpa at the 0.80 and 0.99 levels for the admitted and rejected data here. 

```{r q5}
ttest_gpa_admit_80 = t.test(x=gre_admit$gpa, conf.level=0.80 )
ttest_gpa_admit_99 = t.test(x=gre_admit$gpa, conf.level=0.99 )
ttest_gpa_reject_80 = t.test(x=gre_reject$gpa, conf.level=0.80 )
ttest_gpa_reject_99 = t.test(x=gre_reject$gpa, conf.level=0.99 )
```

I will compare the intervals like I did in Question 3.

At the 0.80 level the CI for rejected is `r ttest_gpa_reject_80$conf.int` and the CI for admitted is `r ttest_gpa_admit_80$conf.int`. Then we check $x_{r} \leq y_{a}$, which gives us `r ttest_gpa_reject_80$conf.int[1] <= ttest_gpa_admit_80$conf.int[2]`, and then we check $y_{r} \leq x_{a}$, which gives us `r ttest_gpa_reject_80$conf.int[2] <= ttest_gpa_admit_80$conf.int[1]`. Because both tests return TRUE, this means that at the 0.80 the CIs do not overlap.   

Now we check if the admitted and rejected overlap at the 0.99 level in the same way. At the 0.99 level the CI for rejected is `r ttest_gpa_reject_99$conf.int` and the CI for admitted is `r ttest_gpa_admit_99$conf.int`. Then we check $x_{r} \leq y_{a}$, which gives us `r ttest_gre_reject_99$conf.int[1] <= ttest_gpa_admit_99$conf.int[2]`, and then we check $y_{r} \leq x_{a}$, which gives us `r ttest_gpa_reject_99$conf.int[2] <= ttest_gpa_admit_99$conf.int[1]`. Because one test returns FALSE, this means that at the 0.99 the CIs do overlap (but only just, you can see that the intervals share the same value on the rejected upper limit and the admitted lower limit).


# Question 6
**Repeat Question 4 for gpa. Make (box-) plots showing the gpa distribution among applicants from different school rankings for the admitted and rejected separately. Again, use ggplot.**

```{r q6, results=T}
#admitted
p_gpa_admit <- ggplot(gre_admit, aes(x=gre_admit$rank, y=gre_admit$gpa)) + geom_boxplot() + stat_summary(fun.y=mean, geom="point", shape=23, size=4) + labs(title="Boxplot of gpa scores \nvs school rankings \nof admitted students", x="school rank", y="gpa scores")
#rejected
p_gpa_reject <- ggplot(gre_reject, aes(x=gre_reject$rank, y=gre_reject$gpa)) + geom_boxplot() + stat_summary(fun.y=mean, geom="point", shape=23, size=4) + labs(title="Boxplot of gpa scores \nvs school rankings \nof rejected students", x="school rank", y="gpa scores")
grid.arrange(p_gpa_admit, p_gpa_reject, ncol=2)
```


# Question 7
**As EDA from the charts and the t-interval calculations you produced, do you observe any potential effects between gre/gpa/rank on admissions? Explain briefly. **

From the t tests we see that the interval (at both the 0.80 and 0.99 level) were higher scores for the admitted students than for the rejected students: 

At the 0.80 level the admitted gre scores are `r ttest_gre_admit_80$conf.int`   
At the 0.80 level the rejected gre scores are `r ttest_gre_reject_80$conf.int` 

At the 0.99 level the admitted gre scores are `r ttest_gre_admit_99$conf.int`   
At the 0.99 level the rejected gre scores are `r ttest_gre_reject_99$conf.int`  

At the 0.80 level the admitted gpa scores are `r ttest_gpa_admit_80$conf.int`   
At the 0.80 level the rejected gpa scores are `r ttest_gpa_reject_80$conf.int` 

At the 0.99 level the admitted gpa scores are `r ttest_gpa_admit_99$conf.int`   
At the 0.99 level the rejected gpa scores are `r ttest_gpa_reject_99$conf.int`

Which would lead one to believe that, at the 0.80 and 0.99 level, higher gre and gpa scores get you accepted. 

If we look at the box plots I have included the mean gre and mean gpa score for the school rank. One might have expected that a higher rank school would have higher average gre and gpa scores but such appears not to be the case. 

```{r q7}
#admitted
admit_rank1 <- subset(gre_admit, rank == 1, select = c("admit","gre","gpa","rank"))
admit_rank2 <- subset(gre_admit, rank == 2, select = c("admit","gre","gpa","rank"))
admit_rank3 <- subset(gre_admit, rank == 3, select = c("admit","gre","gpa","rank"))
admit_rank4 <- subset(gre_admit, rank == 4, select = c("admit","gre","gpa","rank"))
#rejected
reject_rank1 <- subset(gre_reject, rank == 1, select = c("admit","gre","gpa","rank"))
reject_rank2 <- subset(gre_reject, rank == 2, select = c("admit","gre","gpa","rank"))
reject_rank3 <- subset(gre_reject, rank == 3, select = c("admit","gre","gpa","rank"))
reject_rank4 <- subset(gre_reject, rank == 4, select = c("admit","gre","gpa","rank"))
```

I calculate those mean values here, so that they are not just in the box plots.

Rank 1 school admitted mean gre is `r mean(admit_rank1$gre)`  
Rank 2 school admitted mean gre is `r mean(admit_rank2$gre)`   
Rank 3 school admitted mean gre is `r mean(admit_rank3$gre)`    
Rank 4 school admitted mean gre is `r mean(admit_rank4$gre)`  

Rank 1 school rejected mean gre is `r mean(reject_rank1$gre)`  
Rank 2 school rejected mean gre is `r mean(reject_rank2$gre)`  
Rank 3 school rejected mean gre is `r mean(reject_rank3$gre)`  
Rank 4 school rejected mean gre is `r mean(reject_rank4$gre)`  

Rank 1 school admitted mean gpa is `r mean(admit_rank1$gpa)`  
Rank 2 school admitted mean gpa is `r mean(admit_rank2$gpa)`  
Rank 3 school admitted mean gpa is `r mean(admit_rank3$gpa)`  
Rank 4 school admitted mean gpa is `r mean(admit_rank4$gpa)`  

Rank 1 school rejected mean gpa is `r mean(reject_rank1$gpa)`  
Rank 2 school rejected mean gpa is `r mean(reject_rank2$gpa)`  
Rank 3 school rejected mean gpa is `r mean(reject_rank4$gpa)`  
Rank 4 school rejected mean gpa is `r mean(reject_rank4$gpa)` 

We can see from the above data that sometimes Rank 3 schools had higher values for admitted and rejected scores than Rank 2 schools, but Rank 1 schools did have the highest scores and Rank 4 schools did have the lowest scores. There seems to be fluctuation between Rank 2 and Rank 3 scores for who has the higher ones. 








