---
title: "Does The GDP and Population of a Country Affect the Number of Medals They Win and How Does It Affect the Sports They Excel In?"

date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---
The wealth and population of each participating country in the Olympics will be explored in order to build a macro-level insight into how an athlete’s environment may assist or hinder them in winning medals. Here the economic growth and population numbers of each country will be compared in order to ascertain how they may affect the athlete’s ability to win medals.

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T, quietly = T)) stop("Package not found") } }
```

```{r setup2, echo=FALSE, cache=FALSE}
loadPkg("knitr")
loadPkg("rmdformats")

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```
# Datasets

The Olympics dataset obtained previously is merged with datasets containing the Gross Domestic Product (GDP) and population of each country for the years 1960-2019. The datasources were:

Population:

https://knoema.com/UNWPP2019/world-population-prospects-2019

(obtained from worldbank.org)

GDP:

https://data.worldbank.org/indicator/NY.GDP.MKTP.CD

Both datasets had missing data for several of the past decades and for some of the countries that did not participate in the Olympics. Also, for the recent years some of the countries that did participate still had missing data. These were filled by extrapolating the average of the years the preceded/followed the missing data. Furthermore, gross domestic product per capita (GDPpC) was calculated by dividing the GDP over the population. Categorical variables of the data are converted into 'factor' class.

```{r, 1. Read and Save Data, include=FALSE}
#Read data
data <- read.csv('data/athlete_events.csv',stringsAsFactors=FALSE) # affects NA conversion to zero
population <- read.csv('data/pop.csv')
gdp <- read.csv('data/gdp.csv')
loadPkg("dplyr")
loadPkg("ggplot2")
loadPkg("tidyverse")
# install.packages("rmdformats")
# Create Dataframes
olympics_df <- data.frame(data) #Olympics Data
pop_df <- data.frame(population) # Population
gdp_df <- data.frame(gdp) # GDP
```

```{r, 2. Clean Data and Reshape it, include=F}
colnames(pop_df) 
unique(colnames(olympics_df))
unique(olympics_df$Year)
# I will take data from 2000- less missing data and the sports included are the same up to 2016
olympics_mil <- olympics_df %>% filter(Year >= 2000)
unique(olympics_mil$Year)

pop_df_mil <- pop_df %>% select("Country.Code", "X2000", "X2002", "X2004",
                                "X2006", "X2008","X2010","X2012", "X2014", "X2016")
gdp_df_mil <- gdp_df %>% select("Country.Code", "X2000", "X2002", "X2004",
                                "X2006", "X2008","X2010","X2012", "X2014", "X2016")
# Rename columns
colnames(pop_df_mil) = c("NOC", "2000", "2002", "2004", "2006", "2008", "2010",
                         "2012","2014" ,"2016")
colnames(gdp_df_mil) = c("NOC", "2000", "2002", "2004", "2006", "2008", "2010",
                         "2012","2014" ,"2016")
colnames(pop_df_mil)
colnames(gdp_df_mil)
Year <- c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016)
Year

# Reshaping population data by stacking columns and rows differently to be
# consistant with olympics' data
pop_df_mil_t <- data.frame(t(pop_df_mil))
# convert 
pop_df_mil_t[] <- lapply(pop_df_mil_t, as.character)
colnames(pop_df_mil_t) <- pop_df_mil_t[1,] # the first row will be the header
pop_df_mil_t <- pop_df_mil_t[-1, ]          # removing the first row.
pop_df_mil_t <- stack(pop_df_mil_t)
colnames(pop_df_mil_t) = c('Population', 'NOC')
pop_df_mil_t['Population'] <- lapply(pop_df_mil_t['Population'] , as.numeric)
# length(pop_df_mil_t$Population)
pop_df_mil_t$Year <- rep(Year, times = 264)

# Reshaping GDP data by stacking columns and rows differently to be
# consistant with olympics' data
gdp_df_mil_t <- data.frame(t(gdp_df_mil))
gdp_df_mil_t[] <- lapply(gdp_df_mil_t, as.character)
colnames(gdp_df_mil_t) = gdp_df_mil_t[1,] # the first row will be the header
gdp_df_mil_t = gdp_df_mil_t[-1, ]          # removing the first row.
gdp_df_mil_t <- stack(gdp_df_mil_t)
colnames(gdp_df_mil_t) = c('GDP', 'NOC')
gdp_df_mil_t['GDP'] <- lapply(gdp_df_mil_t['GDP'] , as.numeric)
# length(gdp_df_mil_t$GDP)
gdp_df_mil_t$Year <- rep(Year, times = 264)

# Merging popluation and GDP dataframes into one
pop_gdp_mil_df = merge(gdp_df_mil_t, pop_df_mil_t, 
                       by.x=c("NOC", "Year"), 
                       by.y=c("NOC", "Year"))

# checking NAN
# sum(is.na(pop_gdp_mil_df$GDP))
# sum(is.na(pop_gdp_mil_df$Population))

# Replacing missing population values by in the new merged dataframe by interpolating the available years for each country
for(i in seq_along(pop_gdp_mil_df$Population))
  {
  # if missing data is between 2000->2004, interpolate using the two ahead
  if(is.na(pop_gdp_mil_df$Population[i]) & pop_gdp_mil_df$Year[i] <= 2004){
    pop_gdp_mil_df$Population[i] = 
      abs((pop_gdp_mil_df$Population[i+1]/pop_gdp_mil_df$Population[i+2])*
            pop_gdp_mil_df$Population[i+1])
  # if missing data is > 2004, interpolate using the two before
    } else if (is.na(pop_gdp_mil_df$Population[i]) & pop_gdp_mil_df$Year[i] > 2004){
    pop_gdp_mil_df$Population[i] = 
      abs((pop_gdp_mil_df$Population[i-1]/pop_gdp_mil_df$Population[i-2])*
            pop_gdp_mil_df$Population[i-1])
  } else{0
    }
  }
# same as above done for GPD
for(i in seq_along(pop_gdp_mil_df$GDP))
{
  if(is.na(pop_gdp_mil_df$GDP[i]) & pop_gdp_mil_df$Year[i] <= 2004){
    pop_gdp_mil_df$GDP[i] = 
      abs((pop_gdp_mil_df$GDP[i+1]/pop_gdp_mil_df$GDP[i+2])*
            pop_gdp_mil_df$GDP[i+1])
  } else if (is.na(pop_gdp_mil_df$GDP[i]) & pop_gdp_mil_df$Year[i] > 2004){
    pop_gdp_mil_df$GDP[i] = 
      abs((pop_gdp_mil_df$GDP[i-1]/pop_gdp_mil_df$GDP[i-2])*
            pop_gdp_mil_df$GDP[i-1])
  } else{0
  }
}

# Remainder of countries that have NA values do not particpate on the olympics
pop_gdp_mil_df[which(is.na(pop_gdp_mil_df$Population)),]
pop_gdp_mil_df[which(is.na(pop_gdp_mil_df$GDP)),]
# length(unique(pop_gdp_mil_df$NOC))
# unique(olympics_mil$Year)
# setdiff(unique(pop_gdp_mil_df$NOC),unique(olympics_mil$NOC))
olympics_mil <- olympics_mil %>% filter(!is.na(Height))
olympics_mil <- olympics_mil %>% filter(!is.na(Weight))

# replacing NA values for medals with zero
sum(is.na(olympics_mil$Medal))
olympics_mil$Medal[is.na(olympics_mil$Medal)] <- 0
```

```{r, 2 Creating final dataframes, echo=F}
# counting medals won by each country
oly_gpby_NOC_Yr_Mdl <- olympics_mil %>% group_by(NOC, Year, Medal) %>% tally()
oly_gpby_NOC_Yr_Mdl <- data.frame(oly_gpby_NOC_Yr_Mdl)
# counting medals won by each country for every sport
oly_gpby_NOC_Yr_Mdl_sprt <- olympics_mil %>% group_by(Sport, NOC, Year, Medal) %>% count(Medal,)
oly_gpby_NOC_Yr_Mdl_sprt <- data.frame(oly_gpby_NOC_Yr_Mdl_sprt)
oly_gpby_NOC_Yr_Mdl_sprt <- oly_gpby_NOC_Yr_Mdl_sprt %>% filter(Medal!=0)
oly_gpby_NOC_Yr_Mdl_sprt$Sport_fac <- (as.numeric(as.factor(oly_gpby_NOC_Yr_Mdl_sprt$Sport)))

# Merging oly_gpby_NOC_Yr_Mdl df with population and GDP dataframe
pop_gdp_mil_df_oly <- merge(oly_gpby_NOC_Yr_Mdl, pop_gdp_mil_df, 
                       by.x=c("NOC", "Year"), 
                       by.y=c("NOC", "Year"))
# pop_gdp_mil_df_oly[which(is.na(pop_gdp_mil_df_oly$GDP)),]
pop_gdp_mil_df_oly <- pop_gdp_mil_df_oly %>% filter(!is.na(GDP))
pop_gdp_mil_df_oly$GDPpC <- pop_gdp_mil_df_oly$GDP/pop_gdp_mil_df_oly$Population
# sum(pop_gdp_mil_df_oly$n)

# Merging oly_gpby_NOC_Yr_Mdl_sprt df with population and GDP dataframe
pop_gdp_mil_df_sprt <- merge(oly_gpby_NOC_Yr_Mdl_sprt, pop_gdp_mil_df, 
                       by.x=c("NOC", "Year"), 
                       by.y = c("NOC", "Year"), all.x = TRUE)

# pop_gdp_mil_df_sprt[which(is.na(pop_gdp_mil_df_sprt$GDP)),]
pop_gdp_mil_df_sprt <- pop_gdp_mil_df_sprt %>% filter(!is.na(GDP))
pop_gdp_mil_df_sprt$GDPpC <- pop_gdp_mil_df_sprt$GDP/pop_gdp_mil_df_sprt$Population
# sum(pop_gdp_mil_df_sprt$n)
```

# EDA

As a first step to the EDA, for the year 2000 until present, bar charts of the total number of medals won by each country and gross domestic product per capita (GDPpC) were plotted for preliminary comparison.

```{r, 3a. Bar chart: number of medals v country , include=TRUE}
ggplot(pop_gdp_mil_df_oly%>%filter(Medal!=0), aes(x=reorder(NOC, -n), y=n, fill=Medal))+
  geom_bar(stat='identity') + labs(x="Country (NOC)", y="Number of medals (n)")
  ggtitle("Bar chart: number of medals v country")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=5))
```

```{r, 3b Bar chart: number of medals v GDPpC, include=TRUE}
ggplot(pop_gdp_mil_df_oly%>%filter(Medal!=0), aes(x=reorder(NOC, -n), y=(GDPpC)))+
  geom_bar(stat='identity') + labs(x="Country (NOC)", y="Gross Domestic Product (GDPpC) [USD/person")
  ggtitle("Bar chart: number of medals v GDPpC")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=5))
```

Looking at a few randomly picked countries the following is  obsereved:

| Country       | n           | GDPpC rank  |Population  |
|---------------|-------------|-------------|------------|
| Luxembourg    | 0           | 1st         |169th       |
| Norway | 216 | 3rd | 119th |
| USA | 1626 | 7th | 3rd |
| Russia | 933 | 61st | 9th |
| Jordan | 1 | 109th | 88th |
| Ukraine | 161 |119th | 35th |

The table does not directly show a clear and distinct relationship between countries winning medals and their GDP/population. Scatter plots, histograms and pair plots (pairs.panels show bivariate scatter plots below the diagonal, histograms on the diagonal, and the Pearson correlation above the diagonal) are shown below to further investigate the relationship. To simplify the plots, the average GDP, population and GDP per capita are calculated for years of 2000 until present. Also, the outliers were identified using "boxplot.stats" and a plots are split into "with outliers" and "without outliers" are made.

### Relationship between Medals Won and Competing Country's GDP per Capita (year 2000-present)
```{r, 3c. Scatter histogram and correlation pair plots, incldue=TRUE}
scatter_1<-data.frame(unique(pop_gdp_mil_df_oly$NOC))
loadPkg("psych")

for(i in 1: nrow(scatter_1)) {
dummy_set<-subset(pop_gdp_mil_df_oly, pop_gdp_mil_df_oly$NOC==scatter_1$unique.pop_gdp_mil_df_oly.NOC.[i])
dummy_set<-subset(dummy_set, dummy_set$Medal!=0)
scatter_1$totMedals[i]<-sum(dummy_set$n)

dummy_set<-subset(pop_gdp_mil_df_oly, pop_gdp_mil_df_oly$NOC==scatter_1$unique.pop_gdp_mil_df_oly.NOC.[i])
scatter_1$avePop[i]<-mean(dummy_set$Population)

dummy_set<-subset(pop_gdp_mil_df_oly, pop_gdp_mil_df_oly$NOC==scatter_1$unique.pop_gdp_mil_df_oly.NOC.[i])
scatter_1$aveGDP[i]<-mean(dummy_set$GDP)

dummy_set<-subset(pop_gdp_mil_df_oly, pop_gdp_mil_df_oly$NOC==scatter_1$unique.pop_gdp_mil_df_oly.NOC.[i])
scatter_1$aveGDPpC[i]<-mean(dummy_set$GDPpC)
}

eliminateOuts_1 <- scatter_1 %>% filter(!scatter_1$aveGDPpC %in% boxplot.stats(scatter_1$aveGDPpC)$out)


par( mfrow = c( 2, 2 ), oma = c( 0, 0, 2, 0 ) )

plot(x=scatter_1$aveGDPpC, y=scatter_1$totMedals, main="Scatter plot w/ outliers", xlab="Ave. GDPpC [USD pp]", ylab="No. medals won", col = 'red')

plot(x=eliminateOuts_1$aveGDPpC, y=eliminateOuts_1$totMedals, main="Scatter plot w/o outliers", xlab="Ave. GDPpC [USD pp]", ylab="No. medals won", col = 'red')

hist(scatter_1$aveGDPpC, main="Histogram w/ outliers", xlab="Ave. GDPC [USD pp]", col = 'green', breaks =100)

hist(eliminateOuts_1$aveGDPpC, main="Histogram w/o outliers", xlab="Ave. GDPpC [USD pp]", col = 'green', breaks =100)

title("GDPpC and medals won plots", outer=TRUE)

pairs.panels(scatter_1) # skips a position, if needed.

scatter_1_stats= summary(scatter_1)

```

The scatter plots show a positive uptrend. Also the pairplots show a stronger correlation between average GPD and number of medals won when compared to population and GDPpC. The table below summarizes the numbers

| Pearsons cf   | Ave population| Ave GDPpC   |Ave GDP     |
|---------------|---------------|-------------|------------|
| Medals won    | 0.33          | 0.26        |0.85        |

### Relationship Between Sports and Competing Country's GDP per Capita (year 2000-present)

Another relationship to explore is the number of medals won in a particular sport and the GDP, population and GDPpC of the competing countries.The plots below show  scatter and pair plots for three sports. Each exhibiting a different relationship to population, GDP and GDPpC. 

```{r, 3d. Scatter and correlation pair plots final, echo=F, warning=FALSE}
scatter_2 <- pop_gdp_mil_df_sprt %>% group_by(Sport, NOC) %>% tally(n)
scatter_2 <- data.frame(scatter_2)

most_medal_events <- c("Swimming", "Rowing",
"Ice Hockey", "Hockey", "Gymnastics", "Fencing", "Football", "Canoeing", "Basketball", "Wrestling")

for(i in 1: nrow(scatter_2)) {
dummy_set<-subset(pop_gdp_mil_df_sprt, pop_gdp_mil_df_sprt$NOC==scatter_2$NOC[i])
scatter_2$avePop[i]<-mean(unique(dummy_set$Population))

dummy_set<-subset(pop_gdp_mil_df_sprt, pop_gdp_mil_df_sprt$NOC==scatter_2$NOC[i])
scatter_2$aveGDP[i]<-mean(unique(dummy_set$GDP))

dummy_set<-subset(pop_gdp_mil_df_sprt, pop_gdp_mil_df_sprt$NOC==scatter_2$NOC[i])
scatter_2$aveGDPpC[i]<-mean(unique(dummy_set$GDPpC))
}

par( mfrow = c( 1, 3 ), oma = c( 0, 0, 2, 0 ) )

plot(x=subset(scatter_2$aveGDPpC, scatter_2$Sport=="Swimming"), y=subset(scatter_2$n, scatter_2$Sport=="Swimming"), main="Swimming - strong positive correlation", xlab="Average GDP per capita (USD pp)", ylab="Number of medals won", col = 'red')

plot(x=subset(scatter_2$aveGDPpC, scatter_2$Sport=="Boxing"), y=subset(scatter_2$n, scatter_2$Sport=="Boxing"), main="Boxing - neutral relationship", xlab="Average GDP per capita (USD pp)", ylab="Number of medals won", col = 'red')

plot(x=subset(scatter_2$aveGDPpC, scatter_2$Sport=="Canoeing"), y=subset(scatter_2$n, scatter_2$Sport=="Canoeing"), main="Canoeing Swimming - negative correlation", xlab="Average GDP per capita (USD pp)", ylab="Number of medals won", col = 'red')

# ggplot(scatter_2, aes(x= aveGDPpC, y=n, color=Sport))+ geom_point()+
#   ggtitle("Sport scatter plot") +
#   xlab("Dose (mg)") + ylab("Teeth length")

# scatter_2b = scatter_2 %>% filter(Sport == "Equestrianism" | Sport == "Rhythmic Gymnastics" |Sport == "Boxing" )
pairs.panels(scatter_2%>%filter(Sport=="Swimming"), main='Cycling correlation pairplot')
pairs.panels(scatter_2%>%filter(Sport=="Boxing"), main='Boxing correlation pairplot')
pairs.panels(scatter_2%>%filter(Sport=="Canoeing"), main='Canoeing correlation pairplot')
```

The results showed that each sport has a different relationship with when compared with the relevant countries GDPpC. Where swimming has a positive uptrend with GDPpC, boxing is neutral and canoeing has a negative downtrend. The table below summarize the calculated correlation factor for three sports.

| Number of medals won for sport: | Ave population| Ave GDP     |Ave GDPpC   |
|---------------------------------|---------------|-------------|------------|
|  Swimming                       | 0.21          | 0.83        |0.33        |
| Boxing                          | 0.09          | 0.18        |0.04        |
| Canoeing                        | -0.17         | -0.23       |-0.16       |

The Residuals vs Leverage plots below show that none of the datapoints are highly influential against a regression line.

```{r, cooks distance}
swimming <- scatter_2 %>% filter(Sport=="Swimming")
boxing <- scatter_2 %>% filter(Sport=="Boxing")
canoeing <- scatter_2 %>% filter(Sport=="Canoeing")

plot(lm(n ~ aveGDPpC, data=swimming))
plot(lm(n ~ aveGDPpC, data=boxing))
plot(lm(n ~ aveGDPpC, data=canoeing))
# cooks.distance(lm(n ~ aveGDPpC, data=swimming))
# cooks.distance(lm(n ~ aveGDPpC, data=boxing))
# cooks.distance(lm(n ~ aveGDPpC, data=canoeing))
```


# Additional Information
The results of this investigation could be further improved if the data on the percentage of the GDP that gets allocated to the Olympic Committee of each of these countries. Also, the world happiness report, Corruption Perceptions Index, life expectancy and literacy index could all be included to develop a well rounded look into the various factors that my affect an athlete's performance due to the environment they lived in.
