---
title: "preliminary exploration of Olympics data"
author: "Lucia Illari"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: no
    toc_depth: 2
    toc_float: yes
    df_print: paged
    warning: FALSE
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, results = T )
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
```

Loading necessary packages.

```{r packages, include = F}
loadPkg("readr")
loadPkg("tidyr")
loadPkg("ggplot2")
#loadPkg("viridis")
loadPkg("gridExtra")
loadPkg("corrplot")
loadPkg("ggcorrplot")
loadPkg("GGally")
loadPkg("PerformanceAnalytics")
#loadPkg("devtools")
#loadPkg("tm")
#loadPkg("SnowballC")
loadPkg("wordcloud")
loadPkg("RColorBrewer")
```


Loading data.

```{r loadDat, include = F}
athlete_events <- data.frame(read_csv("C:/Users/Squall/Documents/gradschool/classes/Spring 2020/DatSci/Project/athlete_events.csv/athlete_events.csv"))
noc_regions <- data.frame(read_csv("~/gradschool/classes/Spring 2020/DatSci/Project/noc_regions.csv"))
names(athlete_events)
```

Looking at data:

```{r lookAT, results = "markup"}
dim(noc_regions)
summary(athlete_events)
```

Integer encoding the NOC data for possible later usage:

```{r addInt, results = "markup"}
NOC_Int <- c(as.numeric(as.factor(noc_regions$NOC)))
noc_regions$NOC.Int <- NOC_Int
noc_regions
```

Looking at oldest and youngest athletes, finding athletes with NA age:

```{r ages, results = "markup"}
#ages <- athlete_events$Age
#max(athlete_events$Age, na.rm = TRUE)
oldAge <- subset(athlete_events, Age == max(athlete_events$Age, na.rm = TRUE))
youngAge <- subset(athlete_events, Age == min(athlete_events$Age, na.rm = TRUE))
list(oldAge,youngAge)
shyAthletes <- subset(athlete_events, is.na(Age))
head(shyAthletes)
```

Counts of certain columns:

```{r counts, results = "markup"}
loadPkg("dplyr")
sexC <- athlete_events %>% count(Sex)
sexC$percentage <- sexC$n/sum(sexC$n)*100
sexC
ageC <- athlete_events %>% count(Age)
nameC <- athlete_events %>% count(Name)
medalC <- athlete_events %>% count(Medal)
medalC$percentage <- medalC$n/sum(medalC$n)*100
medalC
```

Barchart of medal data:

```{r medalBarPlot, results = "markup"}
medalDat <- subset(athlete_events, !is.na(Medal))
bp1 <- ggplot(athlete_events, aes(x=Medal), color = Team) + geom_bar(aes(y=(..count..)/sum(..count..)))
bp2 <- ggplot(medalDat, aes(x=Medal), color = Team) + geom_bar(aes(y=(..count..)/sum(..count..)))

grid.arrange(bp1, bp2, nrow = 1)
```

Integer encoding data so that it can be include in checking correlation:

```{r integerEnc, results = "markup"}
mDat <- medalDat %>% mutate(Medal = replace(Medal, Medal == "Gold", 1))
mDat <- mDat %>% mutate(Medal = replace(Medal, Medal == "Silver", 2))
mDat <- mDat %>% mutate(Medal = replace(Medal, Medal == "Bronze", 3))
mDat <- mDat %>% mutate(Sex = replace(Sex, Sex == "F", 1))
mDat <- mDat %>% mutate(Sex = replace(Sex, Sex == "M", 2))
mNOCi <- c(as.numeric(as.factor(mDat$NOC)))
mSPi <- c(as.numeric(as.factor(mDat$Sport)))
mDat$NOC.Int <- mNOCi
mDat$Sport.Int <- mSPi
head(mDat)
```

Columns that I want to keep for correlation plot:

```{r keep, results = "markup"}
keeps <- c("ID","Sex","Age","Height","Weight","NOC.Int","Sport.Int","Medal")
dmDat <- mDat[ , keeps, drop = FALSE]
dmDat[,] <- sapply(dmDat[,], as.numeric)
head(dmDat)
```

Correlation plot:

```{r corrPlots, results = "markup"}
loadPkg("digest")
corr = cor(dmDat, use="pairwise.complete.obs") 
c1 <- corrplot.mixed(corr, title="Correlation Matrix", mar=c(0,0,1,0))
c2 <- corrplot(corr, method="pie", type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
#h1 <- heatmap(x=corr, symm=TRUE)

p.mat <- cor_pmat(dmDat)
gcorr1 <- ggcorrplot(corr, method = "circle", hc.order = TRUE)
gcorr2 <- ggcorrplot(corr, method = "square", hc.order = TRUE)
grid.arrange(gcorr1, gcorr2, nrow = 1)
```

Essentially very similar to above, just with package PerformanceAnalytics:

```{r PerfAn, results = "markup"}
chart.Correlation(dmDat, histogram=TRUE, pch=19)
```

In the above plot:
* The distribution of each variable is shown on the diagonal.
* On the bottom of the diagonal : the bivariate scatter plots with a fitted line are displayed
* On the top of the diagonal : the value of the correlation plus the significance level as stars
* Each significance level is associated to a symbol : p-values(0, 0.001, 0.01, 0.05, 0.1, 1) <=> symbols(“***”, “**”, “*”, “.”, " “)

Starting to look at names for analysis:

```{r}
athlete_events[order(athlete_events$Year),][match(unique(athlete_events[order(athlete_events$Year),]$Sex), athlete_events[order(athlete_events$Year),]$Sex),]
```


```{r first_inst, include = F}
loadPkg("stringr")
first.instance <- athlete_events[match(unique(athlete_events$ID), athlete_events$ID),]
last_names <- word(first.instance$Name,-1)
first_names <- word(first.instance$Name,1)
first.instance$First.Name <- first_names
first.instance$Last.Name <- last_names

first.instance.F <- subset(first.instance, first.instance$Sex == "F")
first.instance.M <- subset(first.instance, first.instance$Sex == "M")
```

Using that list of first instances of names as a way to generate global heat plot:

```{r}
getwd()
US_cities_data <- read_csv("US_cities_data.csv")
US_cities_data
US_cities_data100 <- US_cities_data[1:100,]
US_cities_data100[order(-US_cities_data100$num.athletes),]
US_cities_data100 <- US_cities_data100[seq(dim(US_cities_data100)[1],1),]
US_cities_data100
```

```{r}
loadPkg("usmap")
loadPkg("ggplot2")
loadPkg("ggmap")
loadPkg("maps")

US_cities_data100 <- cbind(geocode(as.character(US_cities_data100$name)), US_cities_data100)
```

```{r}
#LA <- map_data("state", region="louisiana")
#salesCalls <- data.frame(State=rep("louisiana",5), City=c("Baton Rouge", "New Orleans", "Shreveport", "Lafayette", "Mandeville"), Calls=c(10,5,8,13,2))

map<-get_map(location='united states', zoom=4, maptype = "watercolor", source='stamen')
pdf("US_map.pdf",width=15,height=15,pointsize=15)
print(ggmap(map) + geom_point(data = US_cities_data100, aes(x=lon, y=lat, show_guide = TRUE, color=num.athletes), alpha=.85, na.rm = T, size = 10) + scale_color_gradient(low="olivedrab1", high="black") + labs(title="Heat map of US birth cities of olympic athletes")+theme_classic(base_size=24))
dev.off()

#USp <- plot_usmap(regions = "counties") + labs(title = "US Counties") + theme(panel.background = element_rect(color = "black", fill = "lightblue"))
```


```{r worldmap, results = "markup"}
loadPkg("tidyverse")
loadPkg("rvest")
loadPkg("magrittr")
loadPkg("ggmap")

first.inst.NOC <- first.instance %>% count(NOC)
#first.inst.NOC <- athlete_events %>% count(NOC)
first.inst.NOC <- first.inst.NOC[order(first.inst.NOC$NOC),]
names(first.inst.NOC)[names(first.inst.NOC) == "n"] <- "rank"
first.inst.NOC$country = noc_regions$region
first.inst.NOC <- first.inst.NOC[order(-first.inst.NOC$rank),]
head(first.inst.NOC,20)

map.world <- map_data("world")
map.world_joined <- left_join(map.world, first.inst.NOC, by = c('region' = 'country'))
map.world_joined <- map.world_joined %>% mutate(fill_flg = ifelse(is.na(rank),F,T))

register_google(key = "")
df.country_points <- data.frame(country = c("Singapore","luxembourg"),stringsAsFactors = F)
geocode.country_points <- geocode(df.country_points$country)
df.country_points <- cbind(df.country_points,geocode.country_points)

w <- ggplot()
w <- w + geom_polygon( data=map.world_joined, aes(x=long, y=lat, group=group), color="black", fill="lightblue" )

w <- w + geom_polygon( data=map.world_joined, aes(x=long, y=lat, group=group, fill = rank), color="black") 

pdf("global_map.pdf",width=25,height=15,pointsize=50)
print(w <- w + scale_fill_continuous(name="counts", low = "goldenrod1", high = "firebrick4",limits = c(0,19400), breaks=c(0, 970, 1940, 2910, 3880, 4850, 5820, 6790, 7760, 8730, 9700, 10670, 11640, 12610, 13580, 14550, 15520, 16490, 17460, 18430, 19400), na.value = "white") + labs(title="Heat map of olympic athletes from around the globe") + theme_gray(base_size=25))
dev.off() 
```


```{r namesCount, include = F}
first.names.count <- first.instance %>% count(First.Name)
first.names.countF <- first.instance.F %>% count(First.Name)
first.names.countM <- first.instance.M %>% count(First.Name)
#first.names.countF[order(first.names.countF$n, decreasing = TRUE),]
#first.names.countM[order(first.names.countM$n, decreasing = TRUE),]
#first.names.count <- data.frame(count(first.instance, vars="First.Name"))
#sort.firstname <- first.names.count[order(first.names.count$freq, decreasing = TRUE),]

last.names.count <- first.instance %>% count(Last.Name)
#sort.lastname <- last.names.count[order(last.names.count$freq, decreasing = TRUE),]
lastname.edit <- subset(last.names.count, !(last.names.count$Last.Name == "Jr.") & !(last.names.count$Last.Name == "Sr.") & !(last.names.count$Last.Name == "III"))
```


```{r wordcloudF, results = "markup"}
pdf("wc1.pdf")
print(wordcloud(words = first.names.count$First.Name, freq = first.names.count$n, min.freq = 1, max.words=100, random.order=FALSE, rot.per=0.35, colors=brewer.pal(8, "Paired")))
dev.off()
pdf("wc2.pdf")
print(wordcloud(words = first.names.countF$First.Name, freq = first.names.countF$n, min.freq = 1, max.words=100, random.order=FALSE, rot.per=0.35, colors=brewer.pal(8, "Paired")))
dev.off()
pdf("wc3.pdf")
print(wordcloud(words = first.names.countM$First.Name, freq = first.names.countM$n, min.freq = 1, max.words=100, random.order=FALSE, rot.per=0.35, colors=brewer.pal(8, "Paired")))
dev.off()
```

```{r wordcloudL, results = "markup"}
pdf("wc4.pdf")
print(wordcloud(words = lastname.edit$Last.Name, freq = lastname.edit$n, min.freq = 1, max.words=100, random.order=FALSE, rot.per=0.35, colors=brewer.pal(8, "Paired")))
dev.off()
```


```{r longest_name, results = "markup"}
val2 = which.max(nchar(athlete_events$Name))
athlete_events[val2,]
```


Looking at height/weight info:

```{r weight_height_info, results = "markup"}
loadPkg("plyr")
weightDat <- subset(first.instance, !is.na(Weight))
heightDat <- subset(first.instance, !is.na(Height))
ageDat <- subset(first.instance, !is.na(Age))
#weightDat <- subset(athlete_events, !is.na(Weight))
#heightDat <- subset(athlete_events, !is.na(Height))
#ageDat <- subset(athlete_events, !is.na(Age))
muW <- ddply(weightDat, "Sex", summarise, grp.mean=mean(Weight), grp.variance=var(Weight), grp.standarddev=sd(Weight))
muH <- ddply(heightDat, "Sex", summarise, grp.mean=mean(Height), grp.variance=var(Height), grp.standarddev=sd(Height))
muA <- ddply(ageDat, "Sex", summarise, grp.mean=mean(Age), grp.variance=var(Age), grp.standarddev=sd(Age))
muW
muH
muA
```

Standard histograms for height/weight/age:

```{r WHhist, results = "markup"}
p1 <- ggplot(weightDat, aes(x=Weight, color=Sex, fill=Sex)) + geom_histogram(alpha=0.5, binwidth = 5, aes(y=..density..), position="dodge") + geom_density(alpha=0.6) + geom_vline(data=muW, aes(xintercept=grp.mean, color=Sex), linetype="dashed") +
labs(title="Weight histogram plot",x="Weight(kg)", y = "Density")

p1c <- p1 + scale_color_brewer(palette="Accent")+ scale_fill_brewer(palette="Accent") + theme_minimal()+theme(legend.position="top")

p2 <- ggplot(heightDat, aes(x=Height, color=Sex, fill=Sex)) + geom_histogram(binwidth = 5, alpha=0.5, aes(y=..density..), position="dodge") + geom_density(alpha=0.6) + geom_vline(data=muH, aes(xintercept=grp.mean, color=Sex), linetype="dashed") +
labs(title="Height histogram plot",x="Height(cm)", y = "Density")

p2c <- p2 + scale_color_brewer(palette="Accent")+ scale_fill_brewer(palette="Accent") + 
  theme_minimal()+theme(legend.position="top")

p3 <- ggplot(ageDat, aes(x=Age, color=Sex, fill=Sex)) + geom_histogram(binwidth = 5, alpha=0.5, aes(y=..density..), position="dodge") + geom_density(alpha=0.6) + geom_vline(data=muA, aes(xintercept=grp.mean, color=Sex), linetype="dashed") +
labs(title="Age histogram plot",x="Age(years)", y = "Density")

p3c <- p3 + scale_color_brewer(palette="Accent")+ scale_fill_brewer(palette="Accent") + 
  theme_minimal()+theme(legend.position="top")

grid.arrange(p1c, p2c, p3c, nrow = 2, ncol = 2)
```

```{r outlierKD2}
outlierKD2 <- function(df, var, rm=FALSE, title, colVal) { 
    dt = df
    var_name <- eval(substitute(var),eval(dt))
    na1 <- sum(is.na(var_name))
    m1 <- mean(var_name, na.rm = T)
    par(mfrow=c(2, 3), oma=c(0,0,3,0))
    boxplot(var_name, main=paste("With outliers\nfor", title), ylab = title, col = colVal, horizontal = TRUE)
    hist(var_name, main=paste("With outliers\nfor", title), xlab=title, ylab="frequency", col = colVal, density = 10)
    qqnorm(var_name, main=paste("Q-Q plot with outliers\nfor", title), col = colVal)
    qqline(var_name)
    outlier <- boxplot.stats(var_name)$out
    mo <- mean(outlier)
    var_name <- ifelse(var_name %in% outlier, NA, var_name)
    boxplot(var_name, main=paste("Without outliers\nfor", title), ylab = title, col = colVal, horizontal = TRUE)
    hist(var_name, main=paste("Without outliers\nfor", title), xlab=title, ylab="frequency", col = colVal, density = 10)
    qqnorm(var_name, main=paste("Q-Q plot without outliers\nfor", title), col = colVal)
    qqline(var_name)
    title(paste("Outlier Check for ", title), outer=TRUE)
    na2 <- sum(is.na(var_name))
    #cat("Outliers identified:", na2 - na1, "\n")
    #cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "\n")
    #cat("Mean of the outliers:", round(mo, 2), "\n")
    m2 <- mean(var_name, na.rm = T)
    #cat("Mean without removing outliers:", round(m1, 2), "\n")
    #cat("Mean if we remove outliers:", round(m2, 2), "\n")
    
    #if(rm){
    #    dt[as.character(substitute(var))] <- invisible(var_name)
    #    cat("Outliers successfully removed", "\n")
    #    return(invisible(dt))
    #} else {
    #    cat("Nothing changed", "\n")
    #    return(invisible(df))
    #}
}
```

```{r outlier_stand, results = "markup"}
pdf("weightOut.pdf",width=14,height=4)
print(outlierKD2(weightDat, Weight, TRUE, "weight (kg)", "red"))
dev.off()
pdf("heightOut.pdf",width=14,height=4)
print(outlierKD2(heightDat, Height, TRUE, "height (cm)", "blue"))
dev.off()
pdf("ageOut.pdf",width=14,height=4)
print(outlierKD2(ageDat, Age, TRUE, "age (years)", "green"))
dev.off()

```


```{r outlierGG, include = T}
g1 <- ggplot(weightDat, aes(x=Sex, y=Weight, color = Sex)) + geom_boxplot() + geom_boxplot(outlier.colour="black", outlier.shape=16, outlier.size=1) + labs(title="Weight Boxplot of Olympic Athletes\nWith Outliers", x="sex", y="weight (kg)")
g2 <- ggplot(weightDat) + geom_qq(aes(sample = Weight), color = "red") + geom_qq_line(aes(sample = Weight)) + theme_minimal() + labs(title="Q-Q Plot of Weight of Olympic Athletes\nWith Outliers")

g3 <- ggplot(weightDat.noOut, aes(x=Sex, y=Weight, color = Sex)) + geom_boxplot() + geom_boxplot(outlier.colour="black", outlier.shape=16, outlier.size=1) + labs(title="Weight Boxplot of Olympic Athletes\nWithout Outliers", x="sex", y="weight (kg)")
g4 <- ggplot(weightDat.noOut) + geom_qq(aes(sample = Weight), color = "red") + geom_qq_line(aes(sample = Weight)) + theme_minimal() + labs(title="Q-Q Plot of Weight of Olympic Athletes\nWithout Outliers")

g5 <- ggplot(heightDat, aes(x=Sex, y=Height, color = Sex)) + geom_boxplot() + geom_boxplot(outlier.colour="black", outlier.shape=16, outlier.size=1) + labs(title="Height Boxplot of Olympic Athletes\nWith Outliers", x="sex", y="height (cm)")
g6 <- ggplot(weightDat) + geom_qq(aes(sample = Height), color = "red") + geom_qq_line(aes(sample = Height)) + theme_minimal() + labs(title="Q-Q Plot of Height of Olympic Athletes\nWith Outliers")

g7 <- ggplot(heightDat.noOut, aes(x=Sex, y=Height, color = Sex)) + geom_boxplot() + geom_boxplot(outlier.colour="black", outlier.shape=16, outlier.size=1) + labs(title="Height Boxplot of Olympic Athletes \nWithout Outliers", x="sex", y="height (cm)")
g8 <- ggplot(heightDat.noOut) + geom_qq(aes(sample = Height), color = "red") + geom_qq_line(aes(sample = Height)) + theme_minimal() + labs(title="Q-Q Plot of Height of Olympic Athletes \nWithout Outliers")

grid.arrange(g1,g2,g3,g4, ncol = 2, nrow = 2)
grid.arrange(g5,g6,g7,g8, ncol = 2, nrow = 2)
```





comment comment comment