---
title: "preliminary exploration of Olympics data"
author: "Lucia Illari"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: no
    toc_depth: 2
    toc_float: yes
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = T )
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T)) stop("Package not found") } }
```

Loading necessary packages.

```{r packages, include = F}
loadPkg("readr")
loadPkg("tidyr")
loadPkg("ggplot2")
#loadPkg("viridis")
loadPkg("gridExtra")
loadPkg("corrplot")
loadPkg("ggcorrplot")
loadPkg("GGally")
loadPkg("PerformanceAnalytics")
#loadPkg("devtools")
#loadPkg("tm")
#loadPkg("SnowballC")
loadPkg("wordcloud")
loadPkg("RColorBrewer")
```


Loading data.

```{r loadDat, include = F}
athlete_events <- data.frame(read_csv("C:/Users/Squall/Documents/gradschool/classes/Spring 2020/DatSci/Project/athlete_events.csv/athlete_events.csv"))
noc_regions <- data.frame(read_csv("~/gradschool/classes/Spring 2020/DatSci/Project/noc_regions.csv"))
```

Looking at data:

```{r lookAT, include = T}
dim(noc_regions)
summary(athlete_events)
```

Integer encoding the NOC data for possible later usage:

```{r addInt, include = T}
NOC_Int <- c(as.numeric(as.factor(noc_regions$NOC)))
noc_regions$NOC.Int <- NOC_Int
head(noc_regions)
```

Looking at oldest and youngest athletes, finding athletes with NA age:

```{r ages, include = T}
#ages <- athlete_events$Age
#max(athlete_events$Age, na.rm = TRUE)
oldAge <- subset(athlete_events, Age == max(athlete_events$Age, na.rm = TRUE))
youngAge <- subset(athlete_events, Age == min(athlete_events$Age, na.rm = TRUE))
list(oldAge,youngAge)
shyAthletes <- subset(athlete_events, is.na(Age))
head(shyAthletes)
```

Counts of certain columns:

```{r counts, include = T}
loadPkg("dplyr")
sexC <- athlete_events %>% count(Sex)
ageC <- athlete_events %>% count(Age)
nameC <- athlete_events %>% count(Name)
medalC <- athlete_events %>% count(Medal)
medalC
```

Barchart of medal data:

```{r medalBarPlot, include = T}
medalDat <- subset(athlete_events, !is.na(Medal))
bp1 <- ggplot(athlete_events, aes(x=Medal), color = Team) + geom_bar(aes(y=(..count..)/sum(..count..)))
bp2 <- ggplot(medalDat, aes(x=Medal), color = Team) + geom_bar(aes(y=(..count..)/sum(..count..)))

grid.arrange(bp1, bp2, nrow = 1)
```

Integer encoding data so that it can be include in checking correlation:

```{r integerEnc, include = T}
mDat <- medalDat %>% mutate(Medal = replace(Medal, Medal == "Gold", 1))
mDat <- mDat %>% mutate(Medal = replace(Medal, Medal == "Silver", 2))
mDat <- mDat %>% mutate(Medal = replace(Medal, Medal == "Bronze", 3))
mDat <- mDat %>% mutate(Sex = replace(Sex, Sex == "F", 1))
mDat <- mDat %>% mutate(Sex = replace(Sex, Sex == "M", 2))
mNOCi <- c(as.numeric(as.factor(mDat$NOC)))
mSPi <- c(as.numeric(as.factor(mDat$Sport)))
mDat$NOC.Int <- mNOCi
mDat$Sport.Int <- mSPi
head(mDat)
```

Columns that I want to keep for correlation plot:

```{r keep, include = T}
keeps <- c("ID","Sex","Age","Height","Weight","NOC.Int","Sport.Int","Medal")
dmDat <- mDat[ , keeps, drop = FALSE]
dmDat[,] <- sapply(dmDat[,], as.numeric)
head(dmDat)
```

Correlation plot:

```{r corrPlots, include = T}
loadPkg("digest")
corr = cor(dmDat, use="pairwise.complete.obs") 
c1 <- corrplot.mixed(corr, title="Correlation Matrix", mar=c(0,0,1,0))
c2 <- corrplot(corr, method="pie", type = "upper", order = "hclust", tl.col = "black", tl.srt = 45)
#h1 <- heatmap(x=corr, symm=TRUE)

p.mat <- cor_pmat(dmDat)
gcorr1 <- ggcorrplot(corr, method = "circle", hc.order = TRUE)
gcorr2 <- ggcorrplot(corr, method = "square", hc.order = TRUE)
grid.arrange(gcorr1, gcorr2, nrow = 1)
```

Essentially very similar to above, just with package PerformanceAnalytics:

```{r PerfAn, include = T}
chart.Correlation(dmDat, histogram=TRUE, pch=19)
```

In the above plot:
* The distribution of each variable is shown on the diagonal.
* On the bottom of the diagonal : the bivariate scatter plots with a fitted line are displayed
* On the top of the diagonal : the value of the correlation plus the significance level as stars
* Each significance level is associated to a symbol : p-values(0, 0.001, 0.01, 0.05, 0.1, 1) <=> symbols(“***”, “**”, “*”, “.”, " “)

Starting to look at names for analysis:

```{r first_inst, include = F}
loadPkg("stringr")
first.instance <- athlete_events[match(unique(athlete_events$ID), athlete_events$ID),]
last_names <- word(first.instance$Name,-1)
first_names <- word(first.instance$Name,1)
first.instance$First.Name <- first_names
first.instance$Last.Name <- last_names
```


```{r namesCount, include = F}
first.names.count <- first.instance %>% count(First.Name)
#first.names.count <- data.frame(count(first.instance, vars="First.Name"))
#sort.firstname <- first.names.count[order(first.names.count$freq, decreasing = TRUE),]

last.names.count <- first.instance %>% count(Last.Name)
#sort.lastname <- last.names.count[order(last.names.count$freq, decreasing = TRUE),]
lastname.edit <- subset(last.names.count, !(last.names.count$Last.Name == "Jr.") & !(last.names.count$Last.Name == "Sr.") & !(last.names.count$Last.Name == "III"))
```


```{r wordcloudF, include = T}
wordcloud(words = first.names.count$First.Name, freq = first.names.count$n, min.freq = 190, max.words=100, random.order=FALSE, rot.per=0.35, colors=brewer.pal(8, "Paired"))
```

```{r wordcloudL, include = T}
wordcloud(words = lastname.edit$Last.Name, freq = lastname.edit$n, min.freq = 50, max.words=100, random.order=FALSE, rot.per=0.35, colors=brewer.pal(8, "Paired"))
```


```{r longest_name, include = T}
val2 = which.max(nchar(athlete_events$Name))
athlete_events[val2,]
```


Looking at height/weight info:

```{r weight_height_info, include = T}
loadPkg("plyr")
weightDat <- subset(athlete_events, !is.na(Weight))
heightDat <- subset(athlete_events, !is.na(Height))
muW <- ddply(weightDat, "Sex", summarise, grp.mean=mean(Weight), grp.variance=var(Weight), grp.standarddev=sd(Weight))
muH <- ddply(heightDat, "Sex", summarise, grp.mean=mean(Height), grp.variance=var(Height), grp.standarddev=sd(Height))
muW
muH
```

Standard histograms forheight/weight:

```{r WHhist, include = T}
p1 <- ggplot(weightDat, aes(x=Weight, color=Sex, fill=Sex)) + geom_histogram(alpha=0.5, binwidth = 5, aes(y=..density..), position="dodge") + geom_density(alpha=0.6) + geom_vline(data=muW, aes(xintercept=grp.mean, color=Sex), linetype="dashed") +
labs(title="Weight histogram plot",x="Weight(kg)", y = "Density")

p1c <- p1 + scale_color_brewer(palette="Accent")+ scale_fill_brewer(palette="Accent") + theme_minimal()+theme(legend.position="top")

p2 <- ggplot(heightDat, aes(x=Height, color=Sex, fill=Sex)) + geom_histogram(binwidth = 5, alpha=0.5, aes(y=..density..), position="dodge") + geom_density(alpha=0.6) + geom_vline(data=muH, aes(xintercept=grp.mean, color=Sex), linetype="dashed") +
labs(title="Height histogram plot",x="Height(cm)", y = "Density")

p2c <- p2 + scale_color_brewer(palette="Accent")+ scale_fill_brewer(palette="Accent") + 
  theme_minimal()+theme(legend.position="top")

grid.arrange(p1c, p2c, nrow = 2)
```








comment comment comment