---
title: "Olympics_EDA"
author: "LNT"
date: "1/31/2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T, quietly = T)) stop("Package not found") } }
```

```{r Q1, include=T}
# Load data
# 
athletes <- read.csv("athlete_events.csv") 

colnames(athletes)
loadPkg("dplyr")
loadPkg("ggplot2")

```
# Find the statistics (summary) of the heights, weights, and ages of all Olympic Athletes 

```{r HWA, echo=F} 

dim(athletes)
summary(athletes$Height)
summary(athletes$Weight)
summary(athletes$Age)
```
```{r weight}

hist(athletes$Weight, main="Histogram of Olympians weight", xlab="Weight (kg)", col = 'green' )
boxplot(athletes$Weight, main="Boxplot of Olympians weight", ylab="Weight (cm)", col = 'red' )
qqnorm(athletes$Weight, main="Q-Q plot of Olympians weight") 
qqline(athletes$Weight)
#shapiro.test(athletes$Weight) # Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Weight,pnorm) # Kolmogorov-Smirnov Tests against normal distribution

```

## Height 
```{r Height}
loadPkg("ggplot2")
h <- ggplot(athletes, aes(x=Height, fill=Sex)) 

#historgram of height broken down by gender
h + geom_histogram(color="gray7") + xlab("height (cm)")  + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Height, main="Boxplot of Olympians Height", ylab="Height (cm)", col = 'tomato4' )

qqnorm(athletes$Height, main="Q-Q plot of Olympians weight") 
qqline(athletes$Height)
#shapiro.test(athletes$Height) # Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Height,pnorm) # Kolmogorov-Smirnov Tests against normal distribution


```
## Age 
```{r Age, echo=F}
hist(athletes$Age, main="Histogram of Olympians Age", xlab="Age (years)", col = 'Purple' )
boxplot(athletes$Age, main="Boxplot of Olympians Age", ylab="Age (cm)", col = 'Beige' )
qqnorm(athletes$Age, main="Q-Q plot of Olympians Age") 
qqline(athletes$Age)
#shapiro.test(athletes$Height) # Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Age,pnorm) # Kolmogorov-Smirnov Tests against normal distribution


```


```{r removeWeightOutliers, results='hide'}
athletes = outlierKD2(athletes,Weight,TRUE)
athletes = outlierKD2(athletes,Age,TRUE)
sd.weight <- sd(athletes$Weight, na.rm = TRUE)
mean.weight <- mean(athletes$Weight, na.rm = TRUE)


```
Standard deviation height `r sd.height`
Mean height `r mean.height`
Standard deviation weight `r sd.weight`
Mean Weight `r mean.weight`




## Histogram of Olympic Athletes Weight 
```{r replot_weight}
a1 <- ggplot(athletes, aes(x=Weight, fill=Sex)) 

#historgram of height broken down by gender
a1 + geom_histogram(color="gray7", binwidth = 4) + xlab("Weight (kg)") + facet_wrap( ~ Sex, ncol=2)



boxplot(athletes$Weight, main="Boxplot of Olympians Weight", ylab="Weight (kg)", col = 'orange' )
qqnorm(athletes$Weight, main="Q-Q plot of Olympians Weight") 
qqline(athletes$Weight)

```


## Histogram of Olympic Athletes Age 
```{r replot_Age}
a1 <- ggplot(athletes, aes(x=Age, fill=Sex)) 

#historgram of height broken down by gender
a1 + geom_histogram(color="gray7", binwidth = 1) + xlab("age") + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Age, main="Boxplot of Olympians Age", ylab="Age", col = 'orange' )
qqnorm(athletes$Age, main="Q-Q plot of Olympians Age") 
qqline(athletes$Age)
```

```{r, echo = F}
pnorm(z80height) # pnorm to find the probability
qnorm(pnorm(z80height)) # qnorm to find the z-value, the inverse of pnorm.  In this case, the answer is just z80height
```

```{r, echo=TRUE}
athletes$zheight = scale(athletes$Height)
athletes$zweight = scale(athletes$Weight)
```

```{r medal}
gold_medals <- subset(athletes, Medal == "Gold")
silver_medals <- subset(athletes, Medal == "Silver")
bronze_medals <- subset(athletes, Medal == "Bronze")

gold_medals[["bmi"]] <- (gold_medals$Weight/gold_medals$Height/gold_medals$Height) * 10000

silver_medals[["bmi"]] <- (silver_medals$Weight/silver_medals$Height/silver_medals$Height) * 10000

bronze_medals[["bmi"]] <- (bronze_medals$Weight/bronze_medals$Height/bronze_medals$Height) * 10000
summary(gold_medals$Height)
summary(silver_medals$Height)
summary(bronze_medals$Height)
```

# Weight
## Gold, Silver and Bronze Medal Athletes
```{r Weight, echo=F}
summary(gold_medals$Weight)
summary(silver_medals$Weight)
summary(bronze_medals$Weight)
```

# BMI 
## Gold, Silver and Bronze Medal Athletes
```{r BMI}
summary(gold_medals$bmi)
summary(silver_medals$bmi)
summary(bronze_medals$bmi)
```
```{r loading,echo=F}

```

## Gold Medal Athletes BMI
```{r gold, echo=F}

boxplot(gold_medals$bmi, main="Gold Medal Athletes BMI", ylab="weight (lbs)", col='gold')
hist(gold_medals$bmi, main="Gold Medal BMI", col = 'Gold' ) 

```
## Silver Medal Athletes BMI
```{r silver, echo =F}

boxplot(silver_medals$bmi, main="Silver Medal Athletes BMI", ylab="BMI", col='gray')
hist(silver_medals$bmi, main="Silver Medal Athletes BMI", col = 'gray' ) 
```
## Bronze Medal Athletes BMI
```{r bronze, echo=F}


#historgram of height broken down by gender
ggplot(adult_athletes, aes(x=BMI, fill=BMI_Category))  + geom_histogram(color="gray7") + facet_wrap( ~ Sex, ncol=2) + scale_fill_discrete(name="", labels=c("Underweight", "Normal", "Overweight", "Obese")) 
# + labs(title="BMI of Olympic Athletes \nOver 18 years old", y="Population", x="BMI")


boxplot(bronze_medals$bmi, main="Bronze Medal Athletes BMI", ylab="BMI", col='brown')


hist(bronze_medals$bmi, main="Bronze Medal Athletes BMI", col = 'brown' )
```
```{r scatter_medal, echo=F}

summer_gold <- subset (gold_medals, gold_medals$Season == "Summer" && Sport != "Art Competitions")
winter_gold <- subset (gold_medals, gold_medals$Season == "Winter")


ggplot(data=summer_gold)+
  geom_point(mapping = aes(x=Weight, y=Height, color=Sport))+
  ggtitle("Scatter plot of Summer Gold Medalists Height vs Weight")

ggplot(data=winter_gold)+
  geom_point(mapping = aes(x=Weight, y=Height, color=Sport))
  ggtitle("Scatter plot of Winter Gold Medalists Height vs Weight")

```


#historgram of height broken down by gender
```{r hist_height}
ggplot(female_athletes, aes(x=BMI, fill=BMI_Category))  + geom_histogram(color="gray7") + labs(title="Body Mass Index (BMI) of Female Olympic Athletes Over 18 years old", y="Population", x="BMI") + scale_fill_discrete(name="", labels=c("Underweight", "Normal", "Overweight", "Obese"))

#boxplot of weight over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Height, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of \nOlympic Height \nOver BMI", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 


#boxplot of Height over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Weight, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of \nOlympic Weight \nOver BMI", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 


#boxplot of Age over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Age, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of \nOlympic Age \nOver BMI", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 

```
# How do Olympic athletes with medals measure up in BMI?
```{r BMI_Calculation}

medal_athletes <- subset(athletes, !is.na(Medal))
dim(medal_athletes)
medal_athletes[["BMI"]] <- (medal_athletes$Weight/gold_medals$Height/gold_medals$Height) * 10000


medal_athletes <- medal_athletes %>% mutate(Category = case_when(
  medal_athletes$BMI < 18.5 ~ "Underweight",
  medal_athletes$BMI >= 18.5 & medal_athletes$BMI < 25 ~ "Normal Weight",
  medal_athletes$BMI >= 25 & medal_athletes$BMI < 30 ~ "Overweight",
  TRUE ~ "Obese"
))

ggplot(data=medal_athletes, aes(x=Category, y=Age, fill=Sex))+ geom_boxplot()
                

df <- medal_athletes[,c("ID", "Category")]
df <- df %>% group_by(Category) %>% tally()

barplot(df$n, names.arg=c("Underweight", "Normal", "Overweight", "Obese"), col= "green")

g <- ggplot(medal_athletes, aes(x=Age, y=BMI, fill=Sex))

# Scatterplot
g + geom_point() + 
  geom_smooth(method="lm", se=F) +
  labs( 
       y="Age", 
       x="BMI", 
       title="Scatter plot of Age vs BMI")
```
```{r corr}
loadPkg("ggcorrplot")
df1 <- medal_athletes[,c("Age", "Height", "Weight", "BMI")]
df1 <- na.omit(df1)


# Correlation matrix
corr <- round(cor(df1), 1)

```

```{r medal_athletes_boxplots}
medal_athletes <- subset(adult_athletes, Medal == "Gold" | Medal == "Silver" | Medal == "Bronze")


gold_medals <- subset(adult_athletes, Medal == "Gold")
silver_medals <- subset(adult_athletes, Medal == "Silver")
bronze_medals <- subset(adult_athletes, Medal == "Bronze")

#boxplot BMI for events with most medal
most_medal_events <- c("Swimming", "Rowing",
"Ice Hockey", "Hockey", "Gymnastics", "Fencing", "Football", "Canoeing", "Basketball", "Wrestling")
popular_events <- subset(medal_athletes, Sport %in% most_medal_events)
ggplot(data=popular_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() +
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9)) + labs(title="Boxplot of BMI for Olympic Athletes \nwith a Medal in most popular events") + 
scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))+
scale_color_manual(labels = c("Underweight", "Normal","Overweight"))
```

```{r BMI_Summer_Athletes}
#boxplot BMI for Summer athletes
summer_events <- subset(medal_athletes, Season == "Summer")
ggplot(data=summer_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() +
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9)) + labs(title="Boxplot of BMI for Summer Olympic Athletes \nwith a Medal") + 
scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))+
scale_color_manual(labels = c("Underweight", "Normal","Overweight"))
```
  
```{r BMI_Winter_athletes}
#boxplot BMI for Winter athletes
Winter_events <- subset(medal_athletes, Season == "Winter")
ggplot(data=Winter_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() + 
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9)) + labs(title="Boxplot of BMI for Winter Olympic Athletes \nin Top 10 Events") +scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))
```

```{r t-intervals}
tinterval.bmi.c95.male = t.test(x=adult_athletes$BMI, conf.level=0.95)
tinterval.bmi.c95.female = t.test(x=adult_athletes$BMI, conf.level=0.95)

```

Variable |      Gender          | level 0.99 |
---------|----------------------|------------|
BMI      | Sex  = M | [`r format(tinterval.bmi.c95.male$conf.int,digits=5)`] 
BMI     | Sex = F  | [`r format(tinterval.bmi.c95.female$conf.int,digits=5)`]  

```{r t-test}
t.test(BMI~Sex, data=adult_athletes)
```