---
title: "Olympics_EDA"
author: "LNT"
date: "03/07/2020"
output:
 rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---


```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T, quietly = T)) stop("Package not found") } }
```
```{r load_packages}

loadPkg("ggplot2")
loadPkg("rmdformats")
loadPkg("knitr")
```

```{r setup, echo=FALSE, cache=FALSE}


## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

```{r read_data}
# Load data
# 
athletes <- read.csv("athlete_events.csv") 

colnames(athletes)


```
# Find the statistics (summary) of the heights, weights, and ages of all Olympic Athletes 

```{r structure, echo=F} 
NROW(athletes)
str(athletes)
```
```{r weight}


w <- ggplot(athletes, aes(x=Weight, fill=Sex)) 

#historgram of height broken down by gender
w + geom_histogram(color="gray7") + xlab("weight (kg)")  + facet_wrap( ~ Sex, ncol=2)


#hist(athletes$Weight, main="Histogram of Olympians weight", xlab="Weight (kg)", col = 'green' )
boxplot(athletes$Weight, main="Boxplot of Olympians weight", ylab="Weight (cm)", col = 'red' )
qqnorm(athletes$Weight, main="Q-Q plot of Olympians weight") 
qqline(athletes$Weight)
#shapiro.test(athletes$Weight) # Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Weight,pnorm) # Kolmogorov-Smirnov Tests against normal distribution

```

## Height 
```{r Height}

h <- ggplot(athletes, aes(x=Height, fill=Sex)) 

#histogram of height broken down by gender
h + geom_histogram(color="gray7",aes(y=..density..)) + xlab("height (cm)")  + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Height, main="Boxplot of Olympians Height", ylab="Height (cm)", col = 'tomato4' )

qqnorm(athletes$Height, main="Q-Q plot of Olympians weight") 
qqline(athletes$Height)
#shapiro.test(athletes$Height) # Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Height,pnorm) # Kolmogorov-Smirnov Tests against normal distribution


```
## Age 
```{r Age}
a <-  ggplot(athletes, aes(x=Age, fill=Sex)) 


#histogram of height broken down by gender
a + geom_histogram(color="gray7") + xlab("age")  + facet_wrap( ~ Sex, ncol=2) 

boxplot(athletes$Age, main="Boxplot of Olympians Age", ylab="Age", col = 'Beige' )
qqnorm(athletes$Age, main="Q-Q plot of Olympians Age") 
qqline(athletes$Age)
# Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Age,pnorm) # Kolmogorov-Smirnov Tests against normal distribution


```

```{r outlierKD2}
# Fix outliers
outlierKD2 <- function(df, var, rm=FALSE) { 
    #' Original outlierKD functino by By Klodian Dhana,
    #' https://www.r-bloggers.com/identify-describe-plot-and-remove-the-outliers-from-the-dataset/
    #' Check outliers, and option to remove them, save as a new dataframe. 
    #' @param df The dataframe.
    #' @param var The variable in the dataframe to be checked for outliers
    #' @param rm Boolean. Whether to remove outliers or not.
    #' @return The dataframe with outliers replaced by NA if rm==TRUE, or df if nothing changed
    #' @examples
    #' outlierKD2(mydf, height, FALSE)
    #' mydf = outlierKD2(mydf, height, TRUE)
    #' mydfnew = outlierKD2(mydf, height, TRUE)
    dt = df # duplicate the dataframe for potential alteration
    var_name <- eval(substitute(var),eval(dt))
    na1 <- sum(is.na(var_name))
    m1 <- mean(var_name, na.rm = T)
    par(mfrow=c(2, 2), oma=c(0,0,3,0))
    boxplot(var_name, main="With outliers")
    hist(var_name, main="With outliers", xlab=NA, ylab=NA)
    outlier <- boxplot.stats(var_name)$out
    mo <- mean(outlier)
    var_name <- ifelse(var_name %in% outlier, NA, var_name)
    boxplot(var_name, main="Without outliers")
    hist(var_name, main="Without outliers", xlab=NA, ylab=NA)
    title("Outlier Check", outer=TRUE)
    na2 <- sum(is.na(var_name))
    cat("Outliers identified:", na2 - na1, "\n")
    cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "\n")
    cat("Mean of the outliers:", round(mo, 2), "\n")
    m2 <- mean(var_name, na.rm = T)
    cat("Mean without removing outliers:", round(m1, 2), "\n")
    cat("Mean if we remove outliers:", round(m2, 2), "\n")
    
    # response <- readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
    # if(response == "y" | response == "yes"){
    if(rm){
        dt[as.character(substitute(var))] <- invisible(var_name)
        #assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
        cat("Outliers successfully removed", "\n")
        return(invisible(dt))
    } else {
        cat("Nothing changed", "\n")
        return(invisible(df))
    }
}
```

```{r removeWeightOutliers, results='hide'}
athletes = outlierKD2(athletes,Weight,TRUE)
athletes = outlierKD2(athletes,Age,TRUE)
sd.weight <- sd(athletes$Weight, na.rm = TRUE)
mean.weight <- mean(athletes$Weight, na.rm = TRUE)
sd.height <- sd(athletes$Height, na.rm = TRUE)
mean.height <- mean(athletes$Height, na.rm = TRUE)
sd.age <- sd(athletes$Age, na.rm = TRUE)
mean.age <- mean(athletes$Age, na.rm = TRUE)


round(mean.weight, digits=1)
round(sd.weight, digits=1)

round(mean.height, digits=1)
round(sd.height, digits=1)

round(mean.age)
round(sd.age)

```




## Histogram of Olympic Athletes Weight 
```{r replot_weight}
a1 <- ggplot(athletes, aes(x=Weight, fill=Sex)) 

#histogram of weight broken down by gender
a1 + geom_histogram(color="gray7", binwidth = 4,aes(y=..density..)) + xlab("weight (kg)") + facet_wrap( ~ Sex, ncol=2)



boxplot(athletes$Weight, main="Boxplot of Olympians Weight", ylab="Weight (kg)", col = 'orange' )
qqnorm(athletes$Weight, main="Q-Q plot of Olympians Weight") 
qqline(athletes$Weight)

```


## Histogram of Olympic Athletes Age 
```{r replot_Age}
a1 <- ggplot(athletes, aes(x=Age, fill=Sex)) 

#histogram of height broken down by gender
a1 + geom_histogram(color="gray7", binwidth = 1, aes(y=..density..)) + xlab("age") + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Age, main="Boxplot of Olympians Age", ylab="Age", col = 'orange' )
qqnorm(athletes$Age, main="Q-Q plot of Olympians Age") 
qqline(athletes$Age)
```

```{r, echo = F}
#pnorm(z80height) # pnorm to find the probability
#qnorm(pnorm(z80height)) # qnorm to find the z-value, the inverse of pnorm.  In this case, the answer is just z80height
```

```{r, echo=TRUE}
athletes$zheight = scale(athletes$Height)
athletes$zweight = scale(athletes$Weight)
```

```{r medal}
gold_medals <- subset(athletes, Medal == "Gold")
silver_medals <- subset(athletes, Medal == "Silver")
bronze_medals <- subset(athletes, Medal == "Bronze")

gold_medals[["BMI"]] <- (gold_medals$Weight/gold_medals$Height/gold_medals$Height) * 10000

silver_medals[["BMI"]] <- (silver_medals$Weight/silver_medals$Height/silver_medals$Height) * 10000

bronze_medals[["BMI"]] <- (bronze_medals$Weight/bronze_medals$Height/bronze_medals$Height) * 10000
summary(gold_medals$Height)
summary(silver_medals$Height)
summary(bronze_medals$Height)
```

# Weight
## Gold, Silver and Bronze Medal Athletes
```{r Weight_medals, echo=F}
summary(gold_medals$Weight)
summary(silver_medals$Weight)
summary(bronze_medals$Weight)
```

# BMI 
## Gold, Silver and Bronze Medal Athletes
```{r BMI_medals}
summary(gold_medals$BMI)
summary(silver_medals$BMI)
summary(bronze_medals$BMI)
```
```{r loading,echo=F}

```

## Gold Medal Athletes BMI
```{r gold_BMI, echo=F}

boxplot(gold_medals$BMI, main="Gold Medal Athletes BMI", ylab="weight (lbs)", col='gold')
hist(gold_medals$BMI, main="Gold Medal BMI", col = 'Gold' ) 

```
## Silver Medal Athletes BMI
```{r silver_BMI, echo =F}

boxplot(silver_medals$BMI, main="Silver Medal Athletes BMI", ylab="BMI", col='gray')
hist(silver_medals$BMI, main="Silver Medal Athletes BMI", col = 'gray' ) 
```
## Bronze Medal Athletes BMI
```{r bronze_BMI, echo=F}


#histogram of height broken down by gender
ggplot(bronze_medals, aes(x=BMI, fill=BMI))  + geom_histogram(color="gray7", aes(y=..density..)) + facet_wrap( ~ Sex, ncol=2) + scale_fill_discrete(name="", labels=c("Underweight", "Normal", "Overweight", "Obese")) 
# + labs(title="BMI of Olympic Athletes \nOver 18 years old", y="Population", x="BMI")


boxplot(bronze_medals$BMI, main="Bronze Medal Athletes BMI", ylab="BMI", col='brown')


hist(bronze_medals$BMI, main="Bronze Medal Athletes BMI", col = 'brown' )
```
```{r scatter_medal, echo=F}

summer_gold <- subset (gold_medals, gold_medals$Season == "Summer" && Sport != "Art Competitions")
winter_gold <- subset (gold_medals, gold_medals$Season == "Winter")


ggplot(data=summer_gold)+
  geom_point(mapping = aes(x=Weight, y=Height, color=Sport))+
  ggtitle("Scatter plot of Summer Gold Medalists Height vs Weight")

ggplot(data=winter_gold)+
  geom_point(mapping = aes(x=Weight, y=Height, color=Sport))
  ggtitle("Scatter plot of Winter Gold Medalists Height vs Weight")

```

```{r bmi}
#source: CDC https://www.cdc.gov/healthyweight/assessing/bmi/adult_bmi/index.html
adult_athletes <- athletes %>% filter(Age>=20)
  
#bmi = weight (kg) / [height (m)]^2
adult_athletes[["BMI"]] <- (adult_athletes$Weight/adult_athletes$Height/adult_athletes$Height) * 10000





```

```{r bmi_02, echo=F}

adult_athletes <- adult_athletes  %>% mutate(BMI_Category = case_when(
  adult_athletes $BMI < 18.5 ~ 1,
  adult_athletes $BMI >= 18.5 & adult_athletes $BMI < 25 ~ 2,
  adult_athletes $BMI >= 25 & adult_athletes $BMI < 30 ~ 3,
  TRUE ~ 4
))

adult_athletes$BMI_Category <- as.factor(adult_athletes$BMI_Category)


```

#histogram of height broken down by gender
```{r hist_height}

#boxplot of weight over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Height, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of \nOlympic Height \nOver BMI", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 


#boxplot of Height over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Weight, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of \nOlympic Weight \nOver BMI", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 


#boxplot of Age over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Age, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of \nOlympic Age \nOver BMI", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 

```



# How do Olympic athletes with medals measure up in BMI?


```{r popular_events}
medal_athletes <- subset(adult_athletes, Medal == "Gold" | Medal == "Silver" | Medal == "Bronze")


gold_medals <- subset(adult_athletes, Medal == "Gold")
silver_medals <- subset(adult_athletes, Medal == "Silver")
bronze_medals <- subset(adult_athletes, Medal == "Bronze")



#histogram of height broken down by gender
ggplot(adult_athletes, aes(x=BMI, fill=BMI_Category))  + geom_histogram(color="gray7", aes(y=..density..)) + facet_wrap( ~ Sex, ncol=2) + scale_fill_discrete(name="", labels=c("Underweight", "Normal", "Overweight", "Obese")) 
# + labs(title="BMI of Olympic Athletes \nOver 18 years old", y="Population", x="BMI")


#boxplot BMI for events with most medal
most_medal_events <- c("Swimming", "Rowing",
"Ice Hockey", "Hockey", "Gymnastics", "Fencing", "Football", "Canoeing", "Basketball", "Wrestling")
popular_events <- subset(medal_athletes, Sport %in% most_medal_events)


ggplot(data=popular_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() +
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 5, hjust = 1), axis.text.y = element_text(vjust = 1, size = 5)) + labs(title="Boxplot of BMI for Olympic Athletes \nwith a Medal") + 
scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))+
scale_color_manual(labels = c("Underweight", "Normal","Overweight"))
```

```{r BMI_Summer_Athletes}
#boxplot BMI for Summer athletes
summer_events <- subset(medal_athletes, Season == "Summer")
ggplot(data=summer_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() +
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9)) + labs(title="Boxplot of BMI for Summer Olympic Athletes \nwith a Medal") + 
scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))+
scale_color_manual(labels = c("Underweight", "Normal","Overweight"))
```
  
```{r BMI_Winter_athletes}
#boxplot BMI for Winter athletes
Winter_events <- subset(medal_athletes, Season == "Winter")
ggplot(data=Winter_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() + 
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9)) + labs(title="Boxplot of BMI for Winter Olympic Athletes \nin Top 10 Events") +scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))
```





```{r team_analysis}
loadPkg("kableExtra")
loadPkg("dplyr")
US_team <- subset(popular_events,NOC=="USA")
x <- popular_events %>%
  group_by(Sport) %>% summarise_at(c("Height", "Weight", "BMI"), mean, na.rm = TRUE) %>% arrange(Height)
names(x) <- c("Sport", "Height", "Weight", "BMI")

convert_m_fi <- function(xm){
    xm_ft <- xm * 39.3701 / 12
    xm_ft_int <- floor(xm * 39.3701 / 12)
    xm_inch <- round((xm_ft - xm_ft_int) * 12, 1)
    return(paste(xm_ft_int, "ft", xm_inch, "in"))
}
x_metric <- data.frame(x$Sport, round(x$Height/100, digits = 2), round(x$Weight), round(x$BMI))
names(x_metric)<- c("Sport", "Avg Height (m)", "Avg Weight (kg)", "BMI")

#x_imperial <- data.frame(x$Sport, convert_m_fi(x$Height/100), round(x$Weight*2.2046226218,1), round(x$BMI) ) 
#names(x_imperial) <- c("Sport", " Mean Height", " Mean Weight (lbs)", "Mean BMI")


#kable(x_imperial) %>%
  #kable_styling(bootstrap_options = c("striped", "hover"))

kable(x_metric, caption="US Olympic Team") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

```{r t-intervals}
tinterval.bmi.c95.male = t.test(x=adult_athletes$BMI, conf.level=0.95)
tinterval.bmi.c95.female = t.test(x=adult_athletes$BMI, conf.level=0.95)

```

Variable |      Gender          | level 0.99 |
---------|----------------------|------------|
BMI      | Sex  = M | [`r format(tinterval.bmi.c95.male$conf.int,digits=5)`] 
BMI     | Sex = F  | [`r format(tinterval.bmi.c95.female$conf.int,digits=5)`]  

```{r t-test}
t.test(BMI~Sex, data=adult_athletes)
```