---
title: "Olympics_EDA"
author: "LNT"
date: "03/12/2020"
output:
 rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---


```{r basicfcn, include=F}
loadPkg = function(x) { if (!require(x,character.only=T, quietly =T)) { install.packages(x,dep=T,repos="http://cran.us.r-project.org"); if(!require(x,character.only=T, quietly = T)) stop("Package not found") } }
```
```{r load_packages}

loadPkg("ggplot2")
loadPkg("rmdformats")
loadPkg("knitr")
```

```{r setup, echo=FALSE, cache=FALSE}


## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
options(scientific=T, digits = 3) 
```

```{r read_data}
# Load data
# 
athletes <- read.csv("athlete_events.csv") 

colnames(athletes)


```
# Find the statistics (summary) of the heights, weights, and ages of all Olympic Athletes 

```{r structure, echo=F} 
str(athletes)
```


## Histogram, Boxplot, Q-Q plot of Olympic Athletes Height 
```{r Height}

h <- ggplot(athletes, aes(x=Height, fill=Sex)) 

#histogram of height broken down by gender
h + geom_histogram(color="gray7",aes(y=..density..)) + xlab("height (cm)")  + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Height, main="Boxplot of Olympians Height", ylab="Height (cm)", col = 'tomato4' )

qqnorm(athletes$Height, main="Q-Q plot of Olympians Height") 
qqline(athletes$Height)
#shapiro.test(athletes$Height) # Unable to use Shapiro-Wilk test for normality because sample size must be between 3 and 5000
ks.test(athletes$Height,pnorm) # Kolmogorov-Smirnov Tests against normal distribution


```


```{r outlierKD2}
# Fix outliers
outlierKD2 <- function(df, var, rm=FALSE) { 
    #' Original outlierKD functino by By Klodian Dhana,
    #' https://www.r-bloggers.com/identify-describe-plot-and-remove-the-outliers-from-the-dataset/
    #' Check outliers, and option to remove them, save as a new dataframe. 
    #' @param df The dataframe.
    #' @param var The variable in the dataframe to be checked for outliers
    #' @param rm Boolean. Whether to remove outliers or not.
    #' @return The dataframe with outliers replaced by NA if rm==TRUE, or df if nothing changed
    #' @examples
    #' outlierKD2(mydf, height, FALSE)
    #' mydf = outlierKD2(mydf, height, TRUE)
    #' mydfnew = outlierKD2(mydf, height, TRUE)
    dt = df # duplicate the dataframe for potential alteration
    var_name <- eval(substitute(var),eval(dt))
    na1 <- sum(is.na(var_name))
    m1 <- mean(var_name, na.rm = T)
    par(mfrow=c(2, 2), oma=c(0,0,3,0))
    boxplot(var_name, main="With outliers")
    hist(var_name, main="With outliers", xlab=NA, ylab=NA)
    outlier <- boxplot.stats(var_name)$out
    mo <- mean(outlier)
    var_name <- ifelse(var_name %in% outlier, NA, var_name)
    boxplot(var_name, main="Without outliers")
    hist(var_name, main="Without outliers", xlab=NA, ylab=NA)
    title("Outlier Check", outer=TRUE)
    na2 <- sum(is.na(var_name))
    cat("Outliers identified:", na2 - na1, "\n")
    cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "\n")
    cat("Mean of the outliers:", round(mo, 2), "\n")
    m2 <- mean(var_name, na.rm = T)
    cat("Mean without removing outliers:", round(m1, 2), "\n")
    cat("Mean if we remove outliers:", round(m2, 2), "\n")
    
    # response <- readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
    # if(response == "y" | response == "yes"){
    if(rm){
        dt[as.character(substitute(var))] <- invisible(var_name)
        #assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
        cat("Outliers successfully removed", "\n")
        return(invisible(dt))
    } else {
        cat("Nothing changed", "\n")
        return(invisible(df))
    }
}
```

```{r removeWeightOutliers, results='hide'}
athletes = outlierKD2(athletes,Weight,TRUE)
athletes = outlierKD2(athletes,Age,TRUE)
sd.weight <- sd(athletes$Weight, na.rm = TRUE)
mean.weight <- mean(athletes$Weight, na.rm = TRUE)
sd.height <- sd(athletes$Height, na.rm = TRUE)
mean.height <- mean(athletes$Height, na.rm = TRUE)
sd.age <- sd(athletes$Age, na.rm = TRUE)
mean.age <- mean(athletes$Age, na.rm = TRUE)


round(mean.weight, digits=1)
round(sd.weight, digits=1)

round(mean.height, digits=1)
round(sd.height, digits=1)

round(mean.age)
round(sd.age)

```

## Histogram, Boxplot, Q-Q plot of Olympic Athletes Weight 
```{r replot_weight}
a1 <- ggplot(athletes, aes(x=Weight, fill=Sex)) 

#histogram of weight broken down by gender
a1 + geom_histogram(color="gray7", binwidth = 4,aes(y=..density..)) + xlab("weight (kg)") + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Weight, main="Boxplot of Olympians Weight", ylab="Weight (kg)", col = 'orange' )
qqnorm(athletes$Weight, main="Q-Q plot of Olympians Weight") 
qqline(athletes$Weight)

```


## Histogram, Boxplot, Q-Q plot of Olympic Athletes Age 
```{r replot_Age}
a1 <- ggplot(athletes, aes(x=Age, fill=Sex)) 

#histogram of Age broken down by gender
a1 + geom_histogram(color="gray7", binwidth = 1, aes(y=..density..)) + xlab("age") + facet_wrap( ~ Sex, ncol=2)

boxplot(athletes$Age, main="Boxplot of Olympians Age", ylab="Age", col = 'orange' )
qqnorm(athletes$Age, main="Q-Q plot of Olympians Age") 
qqline(athletes$Age)
```


```{r medal}
gold_medals <- subset(athletes, Medal == "Gold")
silver_medals <- subset(athletes, Medal == "Silver")
bronze_medals <- subset(athletes, Medal == "Bronze")

gold_medals[["BMI"]] <- (gold_medals$Weight/gold_medals$Height/gold_medals$Height) * 10000

silver_medals[["BMI"]] <- (silver_medals$Weight/silver_medals$Height/silver_medals$Height) * 10000

bronze_medals[["BMI"]] <- (bronze_medals$Weight/bronze_medals$Height/bronze_medals$Height) * 10000
```

# Weight
## Gold, Silver and Bronze Medal Athletes
```{r Weight_medals}
summary(gold_medals$Weight)
summary(silver_medals$Weight)
summary(bronze_medals$Weight)
```

# BMI 
## Gold, Silver and Bronze Medal Athletes
```{r BMI_medals}
summary(gold_medals$BMI)
summary(silver_medals$BMI)
summary(bronze_medals$BMI)
```

## Gold Medal Athletes BMI
```{r gold_BMI, echo=F}

boxplot(gold_medals$BMI, main="Gold Medal Athletes BMI", ylab="BMI", col='gold')

gbmi <- ggplot(gold_medals, aes(x=BMI)) 

#histogram of BMI broken down by gender
gbmi + geom_histogram(color="gold", binwidth = 1, aes(y=..density..)) + xlab("BMI") 

```

## Silver Medal Athletes BMI
```{r silver_BMI}

boxplot(silver_medals$BMI, main="Silver Medal Athletes BMI", ylab="BMI", col='gray')

sbmi <- ggplot(silver_medals, aes(x=BMI)) 

#histogram of 
sbmi + geom_histogram(color="blue", binwidth = 1, aes(y=..density..)) + xlab("BMI") 
```

## Bronze Medal Athletes BMI
```{r bronze_BMI, echo=F}
boxplot(bronze_medals$BMI, main="Bronze Medal Athletes BMI", ylab="BMI", col='brown')

#histogram of height broken down by gender
ggplot(bronze_medals, aes(x=BMI, fill=BMI))  + geom_histogram(color="brown", aes(y=..density..)) 

```
```{r scatter_medal}

summer_gold <- subset (gold_medals, gold_medals$Season == "Summer")
winter_gold <- subset (gold_medals, gold_medals$Season == "Winter")


ggplot(data=summer_gold)+
  geom_point(mapping = aes(x=Weight, y=Height, color=Sport))+
  xlab("Weight (kg)") + ylab("Height (cm)") +
  ggtitle("Scatter plot of Summer Gold Medalists Height vs Weight")

ggplot(data=winter_gold)+
  geom_point(mapping = aes(x=Weight, y=Height, color=Sport))+
  xlab("Weight (kg)") + ylab("Height (cm)") +
  ggtitle("Scatter plot of Winter Gold Medalists Height vs Weight")

```

```{r bmi}
loadPkg("dplyr")
#source: CDC https://www.cdc.gov/healthyweight/assessing/bmi/adult_bmi/index.html
adult_athletes <- athletes %>% filter(Age>=20)
  
#bmi = weight (kg) / [height (m)]^2
adult_athletes[["BMI"]] <- (adult_athletes$Weight/adult_athletes$Height/adult_athletes$Height) * 10000





```

```{r bmi_02, echo=F}

adult_athletes <- adult_athletes  %>% mutate(BMI_Category = case_when(
  adult_athletes $BMI < 18.5 ~ 1,
  adult_athletes $BMI >= 18.5 & adult_athletes $BMI < 25 ~ 2,
  adult_athletes $BMI >= 25 & adult_athletes $BMI < 30 ~ 3,
  TRUE ~ 4
))

adult_athletes$BMI_Category <- as.factor(adult_athletes$BMI_Category)


```

## Height vs BMI
```{r hist_height}

#boxplot of Height over BMI Weight Status
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Height, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of Olympic Height Over BMI Weight Status", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese"))  + ylab("Height (cm)")


#boxplot of Weight over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Weight, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of Olympic Weight Over BMI Weight Status", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) + ylab("Weight (kg")


#boxplot of Age over BMI
ggplot(data=adult_athletes, aes(x=BMI_Category, y=Age, fill=Sex)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 12, hjust = 1), axis.text.y = element_text(vjust = 1, size = 12, hjust = 1)) + labs(title="Boxplot of Olympic Age Over BMI Weight Status", x="BMI") + scale_x_discrete(labels=c("Underweight", "Normal", "Overweight", "Obese")) 

```




## Boxplot of All Olympic Athletes (> 20 y) BMI?
```{r popular_events}
medal_athletes <- subset(adult_athletes, Medal == "Gold" | Medal == "Silver" | Medal == "Bronze")


gold_medals <- subset(adult_athletes, Medal == "Gold")
silver_medals <- subset(adult_athletes, Medal == "Silver")
bronze_medals <- subset(adult_athletes, Medal == "Bronze")



#histogram of height broken down by gender
ggplot(adult_athletes, aes(x=BMI, fill=BMI_Category))  + geom_histogram(color="gray7", aes(y=..density..)) + facet_wrap( ~ Sex, ncol=2) + scale_fill_discrete(name="", labels=c("Underweight", "Normal", "Overweight", "Obese")) 
# + labs(title="BMI of Olympic Athletes \nOver 18 years old", y="Population", x="BMI")
```


## Boxplot of BMI vs Sport wih the Most Medals
``` {r bmi_mostmedal}
#boxplot BMI for events with most medal
most_medal_events <- c("Swimming", "Rowing",
"Ice Hockey", "Hockey", "Gymnastics", "Fencing", "Football", "Canoeing", "Basketball", "Wrestling")
popular_events <- subset(medal_athletes, Sport %in% most_medal_events)


ggplot(data=popular_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() +
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 10, hjust = 1), axis.text.y = element_text(vjust = 1, size = 5)) + 
scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))+
scale_color_manual(labels = c("Underweight", "Normal","Overweight"))
```

## Boxplot of BMI vs Sport (Summer) 
```{r BMI_Summer_Athletes}
#boxplot BMI for Summer athletes
summer_events <- subset(medal_athletes, Season == "Summer")
ggplot(data=summer_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() +
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9))+
scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))+
scale_color_manual(labels = c("Underweight", "Normal","Overweight"))
```

## Boxplot of BMI vs Sport (Winter)   
```{r BMI_Winter_athletes}
#boxplot BMI for Winter athletes
Winter_events <- subset(medal_athletes, Season == "Winter")
ggplot(data=Winter_events, aes(x=Sport, y=BMI, fill=Sex)) + geom_boxplot() + 
geom_hline(yintercept=c(18.5), linetype="dashed", color="green")+
geom_hline(yintercept=c(25), linetype="dashed", color="blue")+
geom_hline(yintercept=c(30), linetype="dashed", color="red")+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, size = 9, hjust = 1), axis.text.y = element_text(vjust = 1, size = 9)) +scale_y_continuous(breaks = seq(10, 35,5),limits=c(12, 38))
```


## US Olympic Team 
```{r team_analysis}
loadPkg("kableExtra")

US_team <- subset(popular_events,NOC=="USA")
x <- popular_events %>%
  group_by(Sport) %>% summarise_at(c("Height", "Weight", "BMI"), mean, na.rm = TRUE) %>% arrange(Height)
names(x) <- c("Sport", "Height", "Weight", "BMI")

convert_m_fi <- function(xm){
    xm_ft <- xm * 39.3701 / 12
    xm_ft_int <- floor(xm * 39.3701 / 12)
    xm_inch <- round((xm_ft - xm_ft_int) * 12, 1)
    return(paste(xm_ft_int, "ft", xm_inch, "in"))
}
x_metric <- data.frame(x$Sport, round(x$Height/100, digits = 2), round(x$Weight), round(x$BMI))
names(x_metric)<- c("Sport", "Avg Height (m)", "Avg Weight (kg)", "BMI")

#x_imperial <- data.frame(x$Sport, convert_m_fi(x$Height/100), round(x$Weight*2.2046226218,1), round(x$BMI) ) 
#names(x_imperial) <- c("Sport", " Mean Height", " Mean Weight (lbs)", "Mean BMI")


#kable(x_imperial) %>%
  #kable_styling(bootstrap_options = c("striped", "hover"))

kable(x_metric) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## T-Intervals 95% confidence
```{r t-intervals}
male_athletes <- subset(adult_athletes, Sex == "M")
female_athletes  <- subset(adult_athletes, Sex == "F")

tinterval.bmi.c95.male = t.test(x=male_athletes$BMI, conf.level=0.95)
tinterval.bmi.c95.female = t.test(x=female_athletes$BMI, conf.level=0.95)

```

Variable |      Gender          | level 0.95 |
---------|----------------------|------------|
BMI      | Sex  = M | [`r format(tinterval.bmi.c95.male$conf.int,digits=5)`] 
BMI     | Sex = F  | [`r format(tinterval.bmi.c95.female$conf.int,digits=5)`]  

## T-Test
```{r t-test}
bmi_ttest <- t.test(BMI~Sex, data=adult_athletes)
```

The p-value being small enough to be `r format(bmi_ttest$p.value, digits=5)`, hence as used a two-sided test we can reject the null hypothesis that the means of BMI being equal between male and female athletes. The plots indicate that male Olympian athletes' BMI to be than female Olympian Athletes. Female and male
athletes BMI: `r format(bmi_ttest$estimate)`


Variable |      Gender          | Average |
---------|----------------------|------------|
BMI      | Sex  = F | [`r format(bmi_ttest$estimate[1],digits=3)`] 
BMI     | Sex = M  | [`r format(bmi_ttest$estimate[2],digits=3)`]  

## Anova Test comparing BMI Averages of Athletes From Different Olympic Sporting Events
```{r anova}

anovabmi = aov(BMI ~ Sport, data=popular_events)
summary(anovabmi)
```
The small p-value would confirm that there are significant differences in BMI averages of athletes in each Sport. 